\documentclass[acmsmall, screen, review, anonymous]{acmart}

\usepackage{listings}
\usepackage{subcaption}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{fontawesome5}
\usepackage{mathpartir}
\usepackage{sourcecodepro}
\usepackage{amsthm}


\DeclareFontFamily{U}{min}{}
\DeclareFontShape{U}{min}{m}{n}{<-> udmj30}{}
\newcommand{\yo}{\!\text{\usefont{U}{min}{m}{n}\symbol{'207}}\!}

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]

\lstset{
  basicstyle=\ttfamily\small,
  columns=fullflexible,
  keepspaces=true
}

\lstdefinestyle{oxcaml}{
  keywordstyle=\color{blue},
  emphstyle=\color{violet},
  language=[Objective]Caml,
  morekeywords={effect, exclave},
  escapeinside={(*@}{@*)},
  literate=
    {[]}{$\glob$}{1}
}

\lstdefinestyle{haskell}{
  keywordstyle=\color{blue},
  emphstyle=\color{violet},
  language=Haskell
}

\newcommand{\nat}{\mathbb{N}}
\newcommand{\glob}{\mathop{\Box}}
\newcommand{\nothing}{\text{\textvisiblespace}}
\newcommand{\arr}{\rightarrow}
\newcommand{\cat}{\mathbb{C}}
\newcommand{\catobj}{C}
\newcommand{\op}{\mathtt{op}}
\newcommand{\catop}{\cat^{\op}}
\newcommand{\id}[1]{\mathrm{id}_{#1}}
\newcommand{\setc}{\mathrm{Set}}
\newcommand{\psh}{\widehat{\cat}}
\newcommand{\terminal}{\top}
\newcommand{\initial}{\bot}
\newcommand{\lawc}{\mathrm{Law_{\kappa}}}
\newcommand{\lawfc}{\mathrm{Law}}
\newcommand{\lawn}{\mathrm{Law_{\nat}}}
\newcommand{\lawcop}{\lawc^{\mathtt{op}}}
\newcommand{\pshlawcop}{\widehat{\lawcop}}
\newcommand{\pshlawc}{\widehat{\lawc}}
\newcommand{\freef}{Free}
\newcommand{\constfname}{\Omega}
\newcommand{\constf}[1]{\constfname #1}
\newcommand{\yonedaname}{\yo}
\newcommand{\yoneda}[1]{\yonedaname(#1)}
\newcommand{\homset}[3]{\mathrm{Hom}_{#1}(#2, #3)}
\newcommand{\arrow}[2]{{#2}^{#1}}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\unit}[1]{\eta^{#1}}
\newcommand{\mult}[1]{\mu^{#1}}
\newcommand{\strength}[1]{t^{#1}}
\newcommand{\counit}[1]{\epsilon^{#1}}
\newcommand{\comult}[1]{\delta^{#1}}
\newcommand{\ununit}{\rho}

\newcommand{\reflectname}{\mathord{\Uparrow}}
\newcommand{\reflectraw}[1]{\mathop{\reflectname(#1)}}
\newcommand{\reflect}[2]{\reflectraw{#1}(#2)}
\newcommand{\reflectc}[2]{\mathop{\reflectname_{#1 #2}}}
\newcommand{\performraw}[2]{\mathop{\reflectname(#1(#2))}}
\newcommand{\perform}[3]{\performraw{#1}{#2}(#3)}
\newcommand{\reifyname}{\Downarrow}
\newcommand{\reify}[2]{\mathop{\reifyname}(#1.\,#2)}
\newcommand{\reifyc}[2]{\mathop{\reifyname_{#1 #2}}}
\newcommand{\vars}{\mathcal{X}}
\newcommand{\loc}{\ell}
\newcommand{\locs}{\mathcal{L}}
\newcommand{\outlname}{\mathtt{out}_L}
\newcommand{\outl}[1]{\mathop{\outlname} #1}
\newcommand{\outrname}{\mathtt{out}_R}
\newcommand{\outr}[1]{\mathop{\outrname} #1}
\newcommand{\mon}{\mathcal{T}}
\newcommand{\ind}[1]{W_{#1}}
\newcommand{\indintro}[3]{{#1}(#2,\, #3)}
\newcommand{\retname}{\mathtt{ret}}
\newcommand{\ret}[1]{\mathop{\retname}(#1)}
\newcommand{\indelimname}{\mathtt{rec}}
\newcommand{\indelim}[2]{\mathop{\indelimname} #1 \,\{#2\}}
\newcommand{\indelimboxname}{\indelimname^{\Box}}
\newcommand{\indelimbox}[3]{\mathop{\indelimboxname_{#1}} #2 \,\{#3\}}
\newcommand{\indelimmname}{\indelimname^{m}}
\newcommand{\indelimm}[2]{\mathop{\indelimmname} #1 \,\{#2\}}
\newcommand{\indretcase}[2]{\retname(#1) \Rightarrow #2}
\newcommand{\indcase}[4]{#1(#2, #3) \Rightarrow #4}
\newcommand{\seq}{\,;\,}
\newcommand{\fstname}{\mathop{\pi_1}}
\newcommand{\fst}[1]{\fstname #1}
\newcommand{\sndname}{\mathop{\pi_2}}
\newcommand{\snd}[1]{\sndname #1}
\newcommand{\inlname}{\mathop{\mathtt{in}_L}}
\newcommand{\inl}[1]{\inlname #1}
\newcommand{\inrname}{\mathop{\mathtt{in}_R}}
\newcommand{\inr}[1]{\inrname #1}
\newcommand{\casename}{\mathtt{case}}
\newcommand{\case}[5]{\mathop{\casename} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
\newcommand{\casemname}{\casename^{m}}
\newcommand{\casem}[5]{\mathop{\casemname} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
\newcommand{\caseboxname}{\casename^{\Box}}
\newcommand{\casebox}[5]{\mathop{\caseboxname} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
\newcommand{\types}{\mathrel{:}}
\newcommand{\gtypes}{\mathrel{:_{\Box}}}
\newcommand{\mtypes}{\mathrel{:_{m}}}
\newcommand{\abs}[2]{\lambda#1.\,#2}
\newcommand{\app}[2]{#1\,#2}
\newcommand{\cempty}{\varepsilon}
\newcommand{\ccons}[2]{#1;\,#2}
\newcommand{\lbind}[3]{\ccons{#1}{#2\types#3}}
\newcommand{\mbind}[3]{\ccons{#1}{#2\mtypes#3}}
\newcommand{\gbind}[3]{\ccons{#1}{#2\gtypes#3}}
\newcommand{\cdiv}[1]{#1/\Box}
\newcommand{\cmdiv}[1]{#1/m}
\newcommand{\cldiv}[1]{#1/\nothing}
\newcommand{\bdiv}[1]{\cdiv{\mathop{!_{#1}}}}
\newcommand{\returnname}{\mathop{\mathtt{return}}}
\newcommand{\return}[1]{\returnname\, #1}
\newcommand{\letvname}{\mathop{\mathtt{let}}}
\newcommand{\letv}[3]{\letvname #1 = #2 \mathop{\mathtt{in}} #3}
\newcommand{\gboxname}{\mathop{\mathtt{box}}}
\newcommand{\gbox}[1]{\gboxname #1}
\newcommand{\gunbox}[3]{\mathop{\mathtt{let}} (\gbox{#1}) = #2 \mathop{\mathtt{in}} #3}
\newcommand{\grunname}{\mathtt{run}}
\newcommand{\grun}[1]{\mathop{\grunname}(#1)}
\newcommand{\sub}[2]{[#1\backslash#2]}
\newcommand{\sbst}[3]{#1\sub{#2}{#3}}
\newcommand{\turnv}{\mathrel{\vdash_V}}
\newcommand{\turnc}{\mathrel{\vdash_C}}
\newcommand{\turnstlc}{\mathrel{\vdash_{STLC}}}
\newcommand{\evmap}{\mathop{ev}}
\newcommand{\currymap}{\mathop{\lambda}}
\newcommand{\stepstoname}{\rightsquigarrow}
\newcommand{\stepstov}{\stepstoname_V}
\newcommand{\stepstoc}{\stepstoname_C}
\newcommand{\stepstostlc}{\stepstoname_{STLC}}
\newcommand{\stepsnotv}{\not\stepstoname_{V}}
\newcommand{\stepsnotc}{\not\stepstoname_{C}}
\newcommand{\stepsnotstlc}{\not\stepstoname_{STLC}}
\newcommand{\reducestoname}{\stepstoname^{*}}
\newcommand{\reducestov}{\reducestoname_{V}}
\newcommand{\reducestoc}{\reducestoname_{C}}
\newcommand{\reducesnottoc}{\not\reducestoname_{C}}
\newcommand{\reducestostlc}{\reducestoname_{STLC}}
\newcommand{\drightsquigarrow}{\begin{array}{@{}c@{}}\rightsquigarrow \\[-8pt] \rightsquigarrow\end{array}}
\newcommand{\parastepstoname}{\drightsquigarrow}
\newcommand{\parastepstov}{\mathrel{\parastepstoname_{V}}}
\newcommand{\parastepstoc}{\mathrel{\parastepstoname_{C}}}
\newcommand{\parareducestoname}{\parastepstoname^{*}}
\newcommand{\parareducestov}{\mathrel{\parareducestoname_{V}}}
\newcommand{\parareducestoc}{\mathrel{\parareducestoname_{C}}}
\newcommand{\snv}{SN_V}
\newcommand{\snc}{SN_C}
\newcommand{\clocs}[1]{\mathrm{locs}(#1)}
\newcommand{\bound}[1]{\nu(#1)}
\newcommand{\redname}{RED}
\newcommand{\redvname}{\redname_{V}}
\newcommand{\redv}[2]{\redvname{({#1},{#2})}}
\newcommand{\redva}[3]{\redvname{({#1},{#2},{#3})}}
\newcommand{\redcname}{\redname_{C}}
\newcommand{\redc}[2]{\redcname{({#1},{#2})}}
\newcommand{\redca}[3]{\redcname{({#1},{#2},{#3})}}
\newcommand{\redsname}{\redname_{S}}
\newcommand{\reds}[1]{\redsname{({#1})}}
\newcommand{\redsc}[2]{\redsname{({#1},{#2})}}
\newcommand{\mif}{\mathop{\mathrm{if}}}
\newcommand{\mthen}{\mathop{\mathrm{then}}}
\newcommand{\melse}{\mathop{\mathrm{else}}}
\newcommand{\embed}[1]{\mathopen{\lbag}#1\mathclose{\rbag}}
\newcommand{\ctxeq}{\cong_{\mathrm{ctx}}}

\newcommand{\todo}[1]{{\color{red}TODO: #1}}

\title{Effect Reflection}
\author{Leo White}
\affiliation{\institution{Jane Street}\country{UK}}
\date{}
\setcopyright{none}

\begin{document}

\begin{abstract}
  Programming languages increasingly support algebraic effects and
  handlers for writing effectful code in a principled way by temporarily
  extending the language with additional effectful operations. However,
  integrating effects into type systems typically requires specialized
  effect systems that add complexity to the language.

  We introduce \emph{effect reflection}, an alternative presentation of
  algebraic effects and handlers that provides, for any algebraic
  signature $\Sigma$ and type $\tau$, an isomorphism:
  \begin{equation*}
    \ind{\Sigma}(\tau)
    \qquad \cong \qquad
    \yoneda{\Sigma} \arr \glob \tau
  \end{equation*}
  where:
  \begin{itemize}
  \item $\ind{\Sigma}(\tau)$ is the free algebra of $\Sigma$ at $\tau$,
    sometimes called the \emph{freer monad} of $\Sigma$.
  \item $\yoneda{\Sigma}$ is a type representing how to inject operations of
    $\Sigma$ into the current \emph{ambient monad}: the monad describing
  which effects ordinary functions can perform.
  \item$\glob$ is a comonadic \emph{global modality} used to track which values do not
  depend on the ambient monad.
  \end{itemize}
  This isomorphism emerges naturally, as a corollary of the Yoneda
  lemma, from the \emph{Ambient Monad model}: a novel semantic model of
  algebraic effects in terms of presheaves over Lawvere
  theories.

  Effect reflection provides all the expressiveness of typed algebraic
  effects without requiring language-level support for effect systems. We
  demonstrate this by implementing effect reflection as a library in
  OxCaml~\cite{lorenzen2024oxidizing}.
\end{abstract}

\maketitle

\section{Effect reflection and the Ambient Monad model}
Algebraic effects and effect handlers have emerged as a powerful
approach to structured programming with effects. However, integrating
effects into type systems typically requires adding specialized effect
systems. Unfortunately, such effect systems require extensive effect
annotations even for functions that do no use algebraic effects. This
limits the wider adoption of algebraic effects into mainstream
languages.

Recent work, in particular Brachthauser et
al.\cite{brachthauser2022effects} and Tang et al.\cite{tang2025modal},
improves the situation by lowering the annotation burden for functions
not using effects. However, these systems still require specialized type
system support. Can we get the benefits of algebraic effects and
handlers without the need to specialized type system support?

One of the attractive properties of algebraic effects is their elegant
mathematical model. However, these more recent approaches do not come
with such a model. Can we give a simple semantic model of algebraic
effects that justifies our design?

\subsection{Algebraic effects and effect handlers}
Algebraic effects and effect handlers allow the programmer to
temporarily extend the set of available effectful operations.

For example, in OCaml, one can write:
\begin{lstlisting}[style=oxcaml]
    type 'a Effect.t +=
      | Get : int Effect.t
      | Set : int -> unit Effect.t
  
    let handle_state (f : unit -> 'a) : int -> 'a =
      match f () with
      | x -> fun _ -> x
      | effect Get, k -> fun s -> continue k s s
      | effect Set s', k -> fun _ -> continue k () s'
\end{lstlisting}
and for the duration of the \lstinline[style=oxcaml]{f ()} function call
two additional effectful operations are available:
\begin{lstlisting}[style=oxcaml]
    let get () : int = perform Get
    let set (i : int) : unit = perform (Set i)
\end{lstlisting}

Effect handlers have a couple of key benefits compared to using an explicit monad:
\begin{itemize}
\item They allow writing code in ``direct style'' -- i.e. using ordinary
  functions and native arrow types. In particular, higher-order
  functions can automatically support effectful functions.
\item Multiple effect handlers compose with each other automatically in an
  easy-to-understand manner.
\end{itemize}

However, they also have an obvious issue: what happens if you try to
call \lstinline[style=oxcaml]{get} or \lstinline[style=oxcaml]{set}
outside of the \lstinline[style=oxcaml]{handle_state} handler?

In OCaml, the effects are untracked: nothing prevents attempting to
perform an effectful operation that is not currently available,
resulting in an \lstinline[style=oxcaml]{Unhandled} exception being
raised. This is essentially the same approach as unchecked exceptions,
which are a common source of bugs in many languages.

\subsection{Effect systems}

Most algebraic effect systems (\cite{hillerstrom2016liberating},
\cite{leijen2014koka}, \cite{convent2020doo},
\cite{biernacki2019binders}, \cite{brachthauser2020effects},
\cite{xie2022first}) solve the problem of unhandled effects by extending
their type system to explicitly track the effects that a function may
perform.

In such a system, the operations of our integer state example would have
types like:
\begin{lstlisting}[style=oxcaml]
    val handle_state : (unit -[istate | 'e]-> 'a) -> int -['e]-> 'a
    val get : unit -[istate]-> int
    val set : int -[istate]-> unit
\end{lstlisting}
with effect annotations on function arrows describing the monad
implicitly used by those functions, and effect variables used to relate
the monads of different functions.

The details vary between systems, but to avoid excessive user
annotation, such effect systems tend to have subtle inference built on
row polymorphism (\cite{hillerstrom2016liberating},
\cite{leijen2014koka}) or implicit subtyping.

A particular source of complexity in such systems is that they require
\emph{effect polymorphism} to type higher-order functions. For example,
the \lstinline[style=oxcaml]{List.map} function, which maps a function
over a list, must be given a type like:
\begin{lstlisting}[style=oxcaml]
    val map : ('a -['e]-> 'b) -> 'a list -['e]-> 'b list
\end{lstlisting}
which includes effect annotations even though
\lstinline[style=oxcaml]{map} does not use algebraic effects.

\subsection{Extending the Ambient Monad}

Some more recent systems(\cite{brachthauser2020effects},
\cite{brachthauser2022effects}, \cite{tang2025modal}) provide an elegant
alternative to effect polymorphism: they treat effect handlers as
extending the effects available in the \emph{ambient monad} -- the monad
associated with ordinary unannotated arrows. This means that, within a
handler for integer state, all functions are permitted to use integer
state without needing to track this fact in their type. Any effect
annotations on a function's type are requirements for additional effects
over and above what is currently available in the ambient monad.

In such systems the type of \lstinline[style=oxcaml]{List.map} can be:
\begin{lstlisting}[style=oxcaml]
    val map : ('a -> 'b) -> 'a list -> 'b list
\end{lstlisting}
whilst still allowing it to interoperate with functions using algebraic
effects. The types of our integer state operations might be something like:
\begin{lstlisting}[style=oxcaml]
    val get : unit -[istate]-> int
    val set : int -[istate]-> unit
\end{lstlisting}
specifying that an integer state effect must be added to the ambient
monad before they can be used. The type of their handler would be:
\begin{lstlisting}[style=oxcaml]
    val handle_state : (unit -[istate]-> 'a) -> int -> 'a
\end{lstlisting}
indicating that it extends the ambient monad with an additional effect.

Brachth{\"a}user et al.~\cite{brachthauser2020effects} describe this
approach as ``lightweight effect polymorphism'' because it brings much
of the benefits of effect polymorphisms without requiring effect
annotations on all higher-order functions.

However, allowing functions to use any effects available at their
definition site requires restricting them in some way: they must be
prevented from leaving the scope of an effect handler in case they are
implicitly relying on that effect handler.

Existing systems accomplish this by making functions second-class, or
via sophisticated modal effect or capability types that can be used to
track which effect handlers a value may depend on. In this paper we will
show that all that is required is some type-level way to track locality
such as Rust's lifetimes or OxCaml's \lstinline[style=oxcaml]{local}
mode.

\subsection{The Ambient Monad model}

If we are going to make changing the ambient monad the basis of our
algebraic effects then we should build a changeable ambient monad right
into our semantic model. \emph{we should index the meaning of types and
  terms by the ambient monad}. Rather than have the semantics of our
types be sets:
\begin{equation*}
\sem{\tau} \types \setc
\end{equation*}
we should have them be functors from $\lawc$ to $\setc$:
\begin{equation*}
\sem{\tau} \types \lawc \arr \setc
\end{equation*}
where $\lawc$ a suitable category of algebraic theories. For each
object $L \types \lawc$ we write $T_L$ for its corresponding monad.

These functors are presheaves and so they support all the standard
constructions (products, coproducts, exponentials, etc.) needed to
support all the usual type formers (products, sums, arrows, etc.) in the
standard way.

The semantics for function types includes the ambient monad:
\begin{equation*}
  \sem{\tau_1 \arr \tau_2}(L) = \sem{\tau_1}(L) \arr T_L(\sem{\tau_2}(L))
\end{equation*}
This is what allows ordinary functions to perform effects from the
ambient monad.

A few things fall out of this model for free, whose combination suggests
an alternative presentation of algebraic effects.

\subsubsection{Reflecting operations}

The Yoneda embedding of $L \types \lawc$ immediately gives us a suitable
semantics for a family of \emph{effect capability} types $\yoneda{L}$:
\begin{equation*}
\sem{\yoneda{L}} = \homset{\lawc}{L}{\_}
\end{equation*}
Elements of $\sem{\yoneda{L}}(L^\prime)$ represent
the ability to reflect effectful operations from $L$ into the ambient
monad $T_{L^{\prime}}$.

We can expose this ability as a reflection construct: for any
parameterized operation $op \types \tau_p \rightarrowtail \tau_a$ in an
algebraic signature $\Sigma$, we can provide a computation
$\perform{v_h}{op}{v_p}$ that uses the effect capability
$v_h \types \yoneda{\Sigma}$ to perform $op$ parameterized by the value
$v_p \types \tau_p$, returning a value of type $\tau_a$.

These effect capabilities are closely related to \emph{named handlers}
(\cite{xie2022first}) and \emph{lexical effect handlers}
(\cite{biernacki2019binders}). They give you values that can represent
the handlers that are currently in scope.

The Yoneda Lemma applied to the semantics of computations tells us that
functions from $\yoneda{L}$ correspond to the sum of the ambient monad
$T_{L^{\prime}}$ with the monad $T_L$:
\begin{equation}
\label{eq:yoneda}
\sem{\yoneda{L} \arr \tau}(L^{\prime}) \cong T_{L^{\prime}+L}(\sem{\tau}(L^{\prime}+L))
\end{equation}

\subsubsection{The global modality}

$\lawc$ has an initial element $\initial$ corresponding to the identity
monad. We can use this to define a global comonad, closely
related to the purity comonad of Choudhury and
Krishnaswami\cite{choudhury2020recovering}:
\begin{equation*}
\glob(\sem{\tau})(L) = \sem{\tau}(\initial)
\end{equation*}
This amounts to restricting $\tau$ to only the elements which do not
depend on the ambient monad. In particular,
\begin{equation*}
\glob(\sem{\tau_1 \arr \tau_2})(L) \cong \arrow{\sem{\tau_2}(\initial)}{\sem{\tau_1}(\initial)}
\end{equation*}
showing that functions within the global comonad correspond to pure
mathematical functions.

We will use this as the semantics for a global modality $\glob$:
\begin{equation*}
\sem{\glob \tau} = \glob(\sem{\tau})
\end{equation*}

\subsubsection{Reifying terms}

If we restrict our attention to monads that correspond to some algebraic
signature $\Sigma$ with no equations -- i.e. the objects in $\lawc$ that
correspond to free monads -- then we can represent the sum of
$T_{\Sigma}$ with the ambient monad directly as an inductive type
$\ind{\Sigma}$. $\ind{\Sigma}$ is sometimes called the freer
monad~\cite{kiselyov2015freer}.

$\ind{\Sigma}{\tau}$ has a unit constructor $\ret{v_r}$ where
$v_r \types \tau$, and for each parameterized operation
$op \types \tau_p \rightarrowtail \tau_a$ in $\Sigma$ it has a
constructor $\indintro{op}{v_p}{v_k}$ where $v_p \types \tau_p$ and
$v_k \types \glob \tau_a \arr \ind{\Sigma}(\tau)$.

\begin{example}
  $\ind{\Sigma}$ for the signature of the integer state effect in OxCaml
  is equivalent to:
\begin{lstlisting}[style=oxcaml]
    type 'a t =
      | Return of 'a @@ global
      | Get of (int -> 'a t)
      | Set of int * (unit -> 'a t)
\end{lstlisting}
  where \lstinline[style=oxcaml]{@@ global} is the OxCaml syntax
  corresponding to the global modality.
\end{example}

By a theorem of Hyland et al.~\cite{hyland2006combining}, we have that:
\begin{equation*}
\sem{\ind{\Sigma}(\tau)}(L) \cong T_{L+\Sigma}(\sem{\tau}(\initial))
\end{equation*}
which, combined with (\ref{eq:yoneda}), gives us an isomorphism:
\begin{equation*}
\sem{\ind{\Sigma}(\tau)} \cong \sem{\yoneda{\Sigma} \arr \glob \tau}
\end{equation*}

This isomorphism allows us to reify effectful computations as inductive
types, giving the same expressive power as effect handlers. Note that it
can be expressed with only effect capabilities, inductive types and a
global modality.

The forwards direction of the isomorphism is actually already provided
by reflection of operations. We can provide the backwards direction as a
reification construct: for any signature $\Sigma$, we provide the
computation $\reify{\loc}{c}$ that binds $\loc$ to an effect capability
of type $\yoneda{\Sigma}$ in the computation $c \types \glob \tau$ and
reifies the uses of $\loc$ in $c$ as a $\ind{\Sigma}(\tau)$.

\subsection{Locality in OxCaml}
\label{sec:locality}

The Ambient Monad model suggests expressing effect reflection and
reification only really requires a global modality, or more broadly some
type system mechanism for tracking locality. OxCaml has support for
tracking locality in types, which we describe here. For a full treatment
see Lorenzen et al.~\cite{lorenzen2024oxidizing}.

Values in OxCaml are associated with both a type -- which describes how
values can be introduced and eliminated -- and a \emph{mode} -- which
tracks properties related to operations other than introduction and
elimination. Modes are a product of a number of different axes, tracking
a variety of different properties, but here we are only interested in
one of them: \emph{locality}. Ignoring the other axes, a mode is either
\lstinline[style=oxcaml]{local} or \lstinline[style=oxcaml]{global}.

Locality tracks whether a value can escape the current
\emph{region}. By default, the current region ends when a function
returns, so local values cannot be returned from functions.
\begin{lstlisting}[style=oxcaml]
  let leak_local x =
    let (o @ local) = Some x in
    o
  (*@ \textcolor{red}{Error: o} @*)
         (*@ \textcolor{red}{\textasciicircum} @*)
  (*@ \textcolor{red}{This value escapes its region.} @*)
\end{lstlisting}
The \lstinline[style=oxcaml]{exclave} construct allows functions to
return local values.
\begin{lstlisting}[style=oxcaml]
  let return_local x =
    exclave
      (let (o @ local) = Some x in
       o)
\end{lstlisting}
\lstinline[style=oxcaml]{exclave e} can only appear in tail position
within a function.  It ends the function's region early and then
executes the expression \lstinline[style=oxcaml]{e}. Since
\lstinline[style=oxcaml]{e} is executed within the region of the caller
of the function, any local values created within it can be safely
returned to that caller.

Arrow types are labelled with two modes: one for the argument and one
for the return value.
\begin{lstlisting}[style=oxcaml]
  val return_local : 'a @ local -> 'a option @ local
\end{lstlisting}
For convenience, and backwards compatibility with OCaml, either mode can
be omitted, in which case it defaults to
\lstinline[style=oxcaml]{global}.

By default, modes are deep: a local
\lstinline[style=oxcaml]{string list} is a local list of local
strings. However, the depth that a mode applies can be controlled via
\emph{modalities}. Record fields can be annotated with the
\lstinline[style=oxcaml]{global} modality to indicate that the contents
of the field are global even when the surrounding record is local. For
example:
\begin{lstlisting}[style=oxcaml]
  type 'a gbl = { g : 'a @@ global }
\end{lstlisting}
gives a record whose only field has the \lstinline[style=oxcaml]{global}
modality. The contents of the \lstinline[style=oxcaml]{g} field is always
global, for example:
\begin{lstlisting}[style=oxcaml]
  let project_global (r @ local) =
    r.g
\end{lstlisting}
gives a function of type:
\begin{lstlisting}[style=oxcaml]
  val project_global : 'a gbl @ local -> 'a @ global
\end{lstlisting}
Variant constructor arguments can also be annotated with modalities.

Note that, whilst there is a \lstinline[style=oxcaml]{global} modality,
there is no \lstinline[style=oxcaml]{local} modality. It does not make
sense for a value which has the ability to leave a region to contain a
value which doesn't have that ability: when the first value leaves a
region it will implicitly cause the second one to leave that region as
well. This means the \lstinline[style=oxcaml]{global} mode is always
deep.

\subsection{Effect reflection in OxCaml}
\label{sec:reflection-in-oxcaml}

Using OxCaml's support for locality and the untracked algebraic effects
it inherits from OCaml, we can implement effect reflection as a simple
library with no specific compiler support.

Given an algebraic signatures represented as a module with a single type
constructor:
\begin{lstlisting}[style=oxcaml]
    module E : sig type 'a t end
\end{lstlisting}
The library provides $\yoneda{\Sigma}$ as an abstract type
\lstinline[style=oxcaml]{Handler(E).t}, and $\ind{\Sigma}$ as
\lstinline[style=oxcaml]{'a Term(E).t}:
\begin{lstlisting}[style=oxcaml]
    type 'a t = 'a Term(E).t =
      | Return : 'a @@ global -> 'a t
      | Perform : 'r E.t @@ global * ('r, 'a) Continuation(E).t -> 'a t
\end{lstlisting}
For technical reasons (see Section \ref{sec:interface}) we use the type
\lstinline[style=oxcaml]{('r, 'a) Continuation(E).t} in place of
\lstinline[style=oxcaml]{'r -> 'a Term(E).t}. Elements of this type
can be applied using a \lstinline[style=oxcaml]{continue} function.

The reflection and reification operations are provided by the
\lstinline[style=oxcaml]{Reflection} functor as functions
\lstinline[style=oxcaml]{perform} and \lstinline[style=oxcaml]{reify}:
\begin{lstlisting}[style=oxcaml]
    val perform : 'a E.t -> Handler(E).t @ local -> 'a

    val reify : (Handler(E).t @ local -> 'a) @ local -> 'a Term(E).t @ local
\end{lstlisting}

We can use this library to write an effect handler equivalent to the
OCaml one in the opening section:
\begin{lstlisting}[style=oxcaml]
    module State = struct
      type 'a t =
        | Get : int t
        | Set : int -> unit t
    end

    module State_eff = Reflection(State)

    let handle_state (f : Handler(State).t @ local -> 'a) (i : int) : 'a =
      let rec loop (s : int) : 'a Term(State).t -> 'a  = function
        | Return x -> x
        | Perform(Get, k) -> loop s (State_eff.continue k s)
        | Perform(Set s', k) -> loop s' (State_eff.continue k ())
      in
      loop i (State_eff.reify f)
\end{lstlisting}
The handler calls \lstinline[style=oxcaml]{f} and passes it a local value of
type \lstinline[style=oxcaml]{Handler(State).t}. For the duration of the
call to \lstinline[style=oxcaml]{f} the ambient monad is extended with
two additional effectful operations, which can be used via reflection:
\begin{lstlisting}[style=oxcaml]
    let get (h : Handler(State).t @ local) : int =
      State_eff.perform Get h

    let set (h : Handler(State).t @ local) (i : int) : unit =
      State_eff.perform (Set i) h
\end{lstlisting}

The locality of the \lstinline[style=oxcaml]{Handler(State).t} value
passed to \lstinline[style=oxcaml]{f} ensures that these operations can
only be performed when available in the ambient monad. The type system
prevents this handler value from escaping the scope of
\lstinline[style=oxcaml]{f}, which in turn prevents
\lstinline[style=oxcaml]{get} or \lstinline[style=oxcaml]{set} from
being called outside that scope.

\subsubsection{Lightweight effect polymorphism}

The following function uses the state handler to calculate the total of
a list of integers:
\begin{lstlisting}[style=oxcaml]
  let total l =
    handle_state 0 (fun h ->
      List.iter (fun x -> set h (get h + x)) l;
      get h)
\end{lstlisting}
This relies on the locality of the function parameter of
\lstinline[style=oxcaml]{List.iter}:
\begin{lstlisting}[style=oxcaml]
  val iter : ('a -> unit) @ local -> 'a list -> unit
\end{lstlisting}
The \lstinline[style=oxcaml]{local} annotation on the parameter ensures
that any function passed will not escape the current region. This allows
us to pass in a closure that closes over the
\lstinline[style=oxcaml]{local} handler \lstinline[style=oxcaml]{h} and
use it to perform some effects.

This ability for higher-order functions like
\lstinline[style=oxcaml]{List.iter} to be reused with different ambient
monads achieves ``lightweight effect polymorphism''.

\subsection{Contributions}

We make the following contributions:
\begin{itemize}
\item We propose \emph{effect reflecton}, a new presentation of effect
  handlers that ensures all effectful operations are properly handled
  without requiring an effect system.
\item We present a simply-typed lambda calculus with effect
  reflection. We prove that the extended system maintains subject
  reduction, confluence and strong normalization. We prove that the
  ordinary simply-typed lambda calculus embeds into it. (Section
  \ref{sec:language})
\item We propose the Ambient Monad model of effectful computations. We
  use it to give a denotational semantics of our lambda calculus with
  effect reflection.  We prove that this semantics is sound,
  compositional and adequate.  (Section \ref{sec:semantics})
\item We give a simple implementation of effect reflection in OxCaml
  using its support for tracking locality in the type system and the
  untracked algebraic effects it inherits from OCaml. We demonstrate its
  usability with some classic examples algebraic effects. (Section
  \ref{sec:practice}).
\end{itemize}

\section{Simply-typed lambda calculus with effect reflection}
\label{sec:language}

We present a simply-typed lambda calculus extended with effect
reflection. This extension requires adding a separate syntax and typing
judgement for computations, in the style of the Fine-Grained
Call-By-Value lambda calculus\cite{levy2003modelling}, with function
application treated as a computation.

\subsection{The base language}

Our language is simply typed, featuring all the usual type-formers. The
syntax and typing rules for the base language are shown in
Fig.\ref{fig:typing-standard}.
\begin{figure}
\begin{align*}
  \mathrm{variables},&x \in& \vars \\
  \mathrm{types}, \tau ::=\;& \tau \times \tau \mid 1 \mid \tau + \tau
                              \mid \tau \arr \tau \mid \ldots \\
  \mathrm{contexts}, \Gamma ::=\;& \cempty \mid \lbind{\Gamma}{x}{\tau} \mid \ldots \\
  \mathrm{values}, v :=\;& x \mid \abs{x}{c}
                           \mid (v, v) \mid \fst{v} \mid \snd{v} \mid () \\
                           \mid& \inl{v} \mid \inr{v} \mid \case{v}{x}{v}{x}{v}
                           \mid \ldots \\
  \mathrm{computations}, c :=\;& \return{v} \mid \letv{x}{c}{c}
                                 \mid \app{v}{v} \mid \ldots
\end{align*}
\begin{mathpar}
\inferrule{\\}
{\ccons{\lbind{\Gamma_1}{x}{\tau}}{\Gamma_2} \turnv x \types \tau}
\and
\inferrule{\lbind{\Gamma}{x}{\tau_1} \turnc c \types \tau_2}
{\Gamma \turnv \abs{x}{c} \types \tau_1 \arr \tau_2}
\and
\inferrule{\Gamma \turnv v_1 \types \tau_1 \\ \Gamma \turnv v_1 \types \tau_1}
{\Gamma \turnv (v_1, v_2) \types \tau_1 \times \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau_1 \times \tau_2}
{\Gamma \turnv \fst{v} \types \tau_1}
\and
\inferrule{\Gamma \turnv v \types \tau_1 \times \tau_2}
{\Gamma \turnv \snd{v} \types \tau_2}
\and
\inferrule{\\}
{\Gamma \turnv () \types 1}
\and
\inferrule{\Gamma \turnv v \types \tau_1}
{\Gamma \turnv \inl{v} \types \tau_1 + \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau_2}
{\Gamma \turnv \inr{v} \types \tau_1 + \tau_2}
\and
\inferrule{\Gamma \turnv v_1 \types \tau_1 + \tau_2 \\
           \lbind{\Gamma}{x_1}{\tau_1} \turnv v_2 \types \tau_3 \\
           \lbind{\Gamma}{x_2}{\tau_2} \turnv v_3 \types \tau_3}
{\Gamma \turnv \case{v_1}{x_1}{v_2}{x_2}{v_3} \types \tau_3}
\and
\inferrule{\Gamma \turnv v \types \tau}
{\Gamma \turnc \return{v} \types \tau}
\and
\inferrule{\Gamma \turnc c_1 \types \tau_1 \\
           \lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2}
{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2}
\and
\inferrule{\Gamma \turnv v_1 \types \tau_1 \arr \tau_2 \\
           \Gamma \turnv v_2 \types \tau_1}
{\Gamma \turnc v_1 v_2 \types \tau_2}
\end{mathpar}
\caption{Basic syntax and typing rules}
\label{fig:typing-standard}
\end{figure}

For convenience we allow some simple pattern forms in variable binding
positions. We allow $()$ to be used in place of a variable when binding
a unit value, and we allow $(x_1, x_2)$ to be used in place of a single
variable when binding a product value. Unit patterns are encoded as just
an unused variable, whilst product bindings are encoded using a variable
$x_0$ and applying the substitutions
$\sub{x_1}{\fst{x_0}}\sub{x_2}{\snd{x_0}}$ to the body of the binding.

Note that $\casename$ -- the eliminator for sum types -- is a value
rather than a computation. We can define a computation version as:
\begin{align*}
    &\case{v}{x_1}{c_1}{x_2}{c_2} := \\
    &\qquad (\case{v}{x_1}{\abs{()}{c_1}}{x_2}{\abs{()}{c_2}}) \; ()
\end{align*}

\subsection{Locality}

We track locality via a global comonadic modality $\glob$. In our
setting being global means not depending on the ambient monad. For
example, $1 \arr 1$ represents computations that may use operations from
the ambient monad, while $\glob (1 \arr 1)$ contains only pure
computations that ignore the ambient monad.

When using $\gbox{v}$ to build a value with a global type, we can only
use existing values that also have a global type. We achieve this using
a fairly standard approach for modal types: we extend contexts with a
binding form that requires a global type and then we restrict the
context in which we type $v$ to only include such bindings. The new
global binding form is written $\gbind{\Gamma}{x}{\tau}$ and represents
a value in the context of type $\glob \tau$. We create global bindings
in the context using the elimination form: $\gunbox{x}{v_1}{v_2}$. As
with $\casename$, this elimination form is a value, but we can define a
computation version directly in terms of it.

Restricting the context is implemented via an operation on contexts
$\cdiv{\Gamma}$ that removes all local bindings, leaving only the global
ones:
\begin{align*}
  &\cdiv{\cempty} = \cempty \\
  &\cdiv{(\lbind{\Gamma}{x}{\tau})} = \cdiv{\Gamma} \\
  &\cdiv{(\gbind{\Gamma}{x}{\tau})} = \gbind{\cdiv{\Gamma}}{x}{\tau}
\end{align*}

Global computations do not depend on the ambient monad, so we also
include a value form $\grun{c}$, which allows treating a global
computation $c$ as a value.

The global comonad distributes over sum types:
\begin{equation*}
  \glob (\tau_1 + \tau_2)
  \qquad \cong \qquad
  (\glob \tau_1) + (\glob \tau_2)
\end{equation*}
which we realize via $\caseboxname$, a global form of the
$\casename$ construct, which allows you to match on a global sum and get
out global components.

For convenience we allow $(\gbox{x})$ to be used as a pattern in
variable binding positions. These are treated as binding a variable
$x_0$ and prefixing the body of the binding with $\gunbox{x}{x_0}{}$.

We add a modalities category, $m$, to the formal syntax, which is either
blank space or $\glob$, to make it easier to describe rules that are
parametric in the presence or absence of the global modality. As part of
this, the $\cdiv{\Gamma}$ operation is extended to work on modalities
$\cmdiv{\Gamma}$ with $\cldiv{\Gamma} = \Gamma$.

The syntax and typing rules related to locality are in
Fig.\ref{fig:typing-locality}
\begin{figure}
\begin{align*}
  \mathrm{types}, \tau ::=\;& \ldots \mid \glob \tau \mid \ldots \\
  \mathrm{modalities}, m ::=\;& \nothing \mid \glob \\
  \mathrm{contexts}, \Gamma ::=\;& \ldots \mid \gbind{\Gamma}{x}{\tau} \\
  \mathrm{values}, v ::=\;& \ldots \mid \gbox{v} \mid \gunbox{x}{v}{v} \mid
                           \casebox{v}{x}{v}{x}{v} \mid \grun{c} \ldots \\
\end{align*}
\begin{mathpar}
\inferrule{\\}
{\ccons{\mbind{\Gamma_1}{x}{\tau}}{\Gamma_2} \turnv x \types \tau}
\and
\inferrule{\cdiv{\Gamma} \turnv v \types \tau}
{\Gamma \turnv \gbox{v} \types \glob \tau}
\and
\inferrule{\Gamma \turnv v_1 \types \glob \tau_1 \\
           \gbind{\Gamma}{x}{\tau_1} \turnv v_2 \types \tau_2}
{\Gamma \turnv \gunbox{x}{v_1}{v_2} \types \tau_2}
\and
\inferrule{\cmdiv{\Gamma} \turnv v_1 \types \tau_1 + \tau_2 \\
           \mbind{\Gamma}{x_1}{\tau_1} \turnv v_2 \types \tau_3 \\
           \mbind{\Gamma}{x_2}{\tau_2} \turnv v_3 \types \tau_3}
{\Gamma \turnv \casem{v_1}{x_1}{v_2}{x_2}{v_3} \types \tau_3}
\and
\inferrule{\cdiv{\Gamma} \turnc c \types \tau}
{\Gamma \turnv \grun{c} \types \tau}
\end{mathpar}
\caption{Locality syntax and typing rules}
\label{fig:typing-locality}
\end{figure}

\subsection{Signatures and inductive types}

Effect reflection requires, for any signature $\Sigma$ that is being
reflected, an inductive type $\ind{\Sigma}$ to represent the free
algebras of $\Sigma$. $\ind{\Sigma}$ is sometimes called the \emph{freer
  monad} of $\Sigma$~\cite{kiselyov2015freer}. They are a form of
W-type, and in a dependently-typed setting we would probably use W-types
directly.

As is standard for algebraic effects systems, we only support handlers
for algebraic theories without any equations, even though our semantics
is defined in terms of arbitrary theories. We define algebraic
signatures $\Sigma$ as finite sets of operations:
\begin{equation*}
\{ op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1}; \ldots; op_n
\types \tau_{p_n} \rightarrowtail \tau_{a_n} \}
\end{equation*}
An operation type $\tau_p \rightarrowtail \tau_a$ indicates an operation
parameterized by $\tau_p$ and with arity $\tau_a$.

\begin{example}
  The signature for boolean state, which we'll call $\Sigma_{bs}$
  is:
  \begin{equation*}
    \{ \mathrm{get} \types 1 \rightarrowtail (1 + 1);
       \mathrm{set} \types (1 + 1) \rightarrowtail 1 \}
  \end{equation*}
\end{example}

We also allow the sum of two signatures $\Sigma_1 + \Sigma_2$ to be used
as a signature. It has operations:
\begin{align*}
\{ \; &\mathrm{left}(op_{1,1}) \types \tau_{p_{1,1}} \rightarrowtail \tau_{a_{1,1}}
        \seq \ldots \seq
        \mathrm{left}(op_{1, n}) \types \tau_{p_{1, n}} \rightarrowtail \tau_{a_{1, n}}
        \seq \\
      &\mathrm{right}(op_{2, 1}) \types \tau_{p_{2, 1}} \rightarrowtail \tau_{a_{2,1}}
        \seq \ldots \seq
        \mathrm{right}(op_{2, m}) \types \tau_{p_{2, m}} \rightarrowtail \tau_{a_{2, m}}
        \; \}
\end{align*}
where $op_{1,1} \seq \ldots \seq op_{1,n}$ are the operations of
$\Sigma_1$ and $op_{2,1} \seq \ldots \seq op_{2,m}$ are the operations
of $\Sigma_2$.

For each operation $op \types \tau_p \rightarrowtail \tau_a$,
$\ind{\Sigma}(\tau)$ has a constructor $\indintro{op}{v_p}{v_k}$
where $v_p \gtypes \tau_p$ and
$v_k \types \glob \tau_a \arr \ind{\Sigma}(\tau)$. $\ind{\Sigma}(\tau)$
also has a constructor $\ret{v}$ where $v \gtypes \tau$.

The eliminator for $\ind{\Sigma}$ is written
\begin{align*}
\indelim{v_s}
  {&\indretcase{x}{v_e} \seq \\
   &\indcase{op_1}{x_{p_1}}{x_{k_1}}{v_1} \seq \\
   &\ldots \seq \\
   &\indcase{op_n}{x_{p_n}}{x_{k_n}}{v_n}}
\end{align*}
The syntax and typing rules for inductive types are shown in
Fig.\ref{fig:typing-inductive}.

As with $\casename$, $\indelimname$ is defined as a value and we can
define a computation version in terms of it.

The global comonad distributes over inductive types, which we realize
via $\indelimboxname$, a global form of the $\indelimname$ construct,
which allows you to match on a global inductive type and get out global
components.

\begin{figure}
\begin{align*}
  \mathrm{types}, \tau ::=\;& \ldots \mid \ind{\Sigma}(\tau) \mid \ldots \\
  \mathrm{signatures}, \Sigma ::=\;& \{ op \types \tau \rightarrowtail \tau \seq \ldots
                                       \seq op \types \tau \rightarrowtail \tau \}
                                     \mid \Sigma + \Sigma \\
  \mathrm{values}, v :=\;& \ldots \mid \indintro{op}{v}{v}
                           \mid \ret{v} \mid \\
                         & \indelimm{v}
                           {\indretcase{x}{v} \seq
                           \indcase{op}{x}{x}{v} \seq \ldots
                           \seq \indcase{op}{x}{x}{v}} \ldots
\end{align*}
\begin{mathpar}
\inferrule{op \types \tau_1 \rightarrowtail \tau_2 \in \Sigma \\
           \cdiv{\Gamma} \turnv v_1 \types \tau_1 \\
           \Gamma \turnv v_2 \types \glob \tau_2 \arr \ind{\Sigma}(\tau)}
{\Gamma \turnv \indintro{op}{v_1}{v_2} \types \ind{\Sigma}(\tau)}
\and
\inferrule{\cdiv{\Gamma} \turnv v \types \tau}
{\Gamma \turnv \ret{v} \types \ind{\Sigma}(\tau)}
\and
\inferrule{\cmdiv{\Gamma} \turnv v_s \types \ind{\Sigma}(\tau_1) \\
           \Sigma = \{ op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1};
             \ldots; op_n\types \tau_{p_n} \rightarrowtail \tau_{a_n} \} \\
           \gbind{\Gamma}{x_r}{\tau_1} \turnv v_e \types \tau_2 \\
           \mbind{\gbind{\Gamma}{x_{p_1}}{\tau_{p_1}}}{x_{k_1}}{\glob \tau_{a_1} \arr \tau_2} \turnv v_1 \types \tau_2 \\
           \ldots \\
           \mbind{\gbind{\Gamma}{x_{p_n}}{\tau_{p_n}}}{x_{k_n}}{\glob \tau_{a_n} \arr \tau_2} \turnv v_n \types \tau_2}
{\Gamma \turnv \indelimm{v_s}{\indretcase{x_r}{v_e} \seq
                                   \indcase{op_1}{x_{p_1}}{x_{k_1}}{v_1} \seq \ldots \seq
                                   \indcase{op_n}{x_{p_n}}{x_{k_n}}{v_n}} \types \tau_2}
\end{mathpar}
\caption{Inductive type syntax and typing rules}
\label{fig:typing-inductive}
\end{figure}

\begin{example}
  \label{example:negate}
  The following value of type $\ind{\Sigma_{bs}}$ represents a
  reified computation that negates the value of the state:
  \begin{align*}
    &\indintro{get}{()}{\\
      &\quad\abs{(\gbox{x_a})}{\\
          &\qquad \indintro{set}{\grun{\app{\mathit{negate}}{x_a}}}
            {\abs{(\gbox{x_r})}{\ret{x_r}}}}}
  \end{align*}
  where $\mathit{negate}$ is the function that negates a boolean.
\end{example}

\subsection{Effect reflection}

We provides a type $\yoneda{\Sigma}$ for each algebraic signature
$\Sigma$ to represent mappings from the operations of $\Sigma$ into the
current ambient monad. Effect reflection appears as two constructs:
reflection and reification. Reflection $\performraw{v}{op}$ uses a value
$v \types \yoneda{\Sigma}$ to map the operation $op$ from $\Sigma$ into
the ambient monad.  Reification $\reifyname$ extends the ambient monad
with the operations from some $\Sigma$, providing a value of type
$\yoneda{\Sigma}$ for constructing those operations via reflection.
These constructs implement the core isomorphism of effect reflection:
reification can convert functions $\yoneda{\Sigma} \arr \glob \tau$ into
terms $\ind{\Sigma}(\tau)$, while reflection provides the inverse
direction. Their syntax and typing rules are shown in
Fig.\ref{fig:typing-reflection}.

\begin{figure}
\begin{align*}
  \mathrm{locations},&\loc \in& \locs \\
  \mathrm{types}, \tau ::=\;& \ldots \mid \yoneda{\Sigma} \mid \ldots \\
  \mathrm{contexts}, \Gamma ::=\;& \ldots \mid \lbind{\Gamma}{\loc}{\Sigma} \\
  \mathrm{values}, v :=\;& \ldots \mid \loc \mid \outl{v} \mid \outr{v} \\
  \mathrm{computations}, c :=\;& \ldots \perform{v}{op}{v} \mid \reify{\loc}{c}
\end{align*}
\begin{mathpar}
\inferrule{\\}
{\ccons{\lbind{\Gamma_1}{\loc}{\Sigma}}{\Gamma_2} \turnv \loc \types \yoneda{\Sigma}}
\and
\inferrule{\Gamma \turnv v \types \yoneda{\Sigma_1 + \Sigma_2}}
{\Gamma \turnv \outl{v} \types \yoneda{\Sigma_1}}
\and
\inferrule{\Gamma \turnv v \types \yoneda{\Sigma_1 + \Sigma_2}}
{\Gamma \turnv \outr{v} \types \yoneda{\Sigma_2}}
\and
\inferrule{\Gamma \turnv v_1 \types \yoneda{\Sigma} \\
           op \types \tau_1 \rightarrowtail \tau_2 \in \Sigma \\
           \cdiv{\Gamma} \turnv v_2 \types \tau_1}
{\Gamma \turnc \; \perform{v_1}{op}{v_2} \types \glob \tau_2}
\and
\inferrule{\lbind{\Gamma}{\loc}{\Sigma} \turnc c \types \glob \tau}
{\Gamma \turnc \; \reify{\loc}{c} \types \ind{\Sigma}(\tau)}
\end{mathpar}
\caption{Effect reflection syntax and typing rules}
\label{fig:typing-reflection}
\end{figure}

We introduce a separate set of variables $\loc$ for the bindings
introduced by reification. This makes it easier to define the desired
reduction rules and normal forms. We give them their own binding form
$\loc \types \Sigma$ in the context, which are also removed by the
$\cdiv{}$ operation

Note that we treat reflection of individual operations
$\performraw{v}{op}$ as primitive. We can define reflection of
terms $\reflectraw{v}$ in terms of $\performraw{v}{op}$:
\begin{align*}
    \reflect{v}{c} := \quad &\letv{x_w}{c}{\\
        &\indelim{x_w}{ \\
        &\qquad \indretcase{x_r}{\return{(\gbox{x_r}})} \seq \\
        &\qquad \indcase{op_1}{x_{p_1}}{x_{k_1}}
          {\letv{x_{a_1}}{\perform{v}{op_1}{x_{p_1}}}{\app{x_{k_1}}{x_{a_1}}}} \seq \\
        &\qquad \ldots \seq \\
        &\qquad \indcase{op_n}{x_{p_n}}{x_{k_n}}
          {\letv{x_{a_n}}{\perform{v}{op_n}{x_{p_n}}}{\app{x_{k_n}}{x_{a_n}}}}}}
\end{align*}
which has this admissible typing rule:
\begin{mathpar}
\inferrule{\Gamma \turnv v \types \yoneda{\Sigma} \\
           \Gamma \turnv c \types \ind{\Sigma}(\tau)}
{\Gamma \turnc \; \reflect{v}{c} \types \glob \tau}
\end{mathpar}

\begin{example}
  The value from Example~\ref{example:negate} can also be constructed
  via reflection:
  \begin{align*}
    &\reify{\loc}{\\
      &\qquad \letv{(\gbox{x})}{\perform{\loc}{get}{}}{\\
      &\qquad \perform{\loc}{set}{\grun{\app{\mathit{negate}}{x}}}}}
  \end{align*}
  where $\mathit{negate}$ is the function that negates a boolean.
\end{example}

We also provide projections $\outl$ and $\outr$ that decompose a mapping
for $\Sigma_1 + \Sigma_2$ into mappings for $\Sigma_1$ and $\Sigma_2$
respectively. This is actually crucial to obtain the full expressivity
of effect handlers. Without it the language would be equivalent to the
more restrictive ``scoped resumptions'' form of effect handlers proposed
by Xie et al.\cite{xie2020effect}.

\subsection{Operational semantics}
\label{sec:operational}

We define the operational semantics as a pair of mutually defined
reduction relations $\stepstov$ and $\stepstoc$ for values and
computations respectively shown in Fig.\ref{fig:reduction}, omitting the
obvious congruence rules. We write $\reducestov$ and $\reducestoc$ for
the reflexive transitive closures of these relations. The definitions of
normal forms for values and computations are given in
Fig.\ref{fig:normal}.

\begin{figure}
\begin{align*}
\fst{(v_1, v_2)} &\stepstov v_1 \\
\snd{(v_1, v_2)} &\stepstov v_2 \\
\casem{(\inl{v_1})}{x_1}{v_2}{x_2}{v_3} &\stepstov \sbst{v_2}{x_1}{v_1} \\
\casem{(\inr{v_1})}{x_1}{v_2}{x_2}{v_3} &\stepstov \sbst{v_3}{x_2}{v_1} \\
\gunbox{x}{\gbox{v_1}}{v_2} &\stepstov \sbst{v_2}{x}{v_1} \\
\grun{\return{v}} &\stepstov v \\
\indelimm{(\ret{v_r})}{\indretcase{x}{v_e} \seq \ldots} &\stepstov \sbst{v_e}{x}{v_r} \\
\indelimm{(\indintro{op}{v_1}{v_2})}{H_1 \seq \indcase{op}{x_1}{x_2}{v_3} \seq H_2} &\stepstov \\
  \sbst{\sbst{v_3}{x_1}{v_1}}{x_2}{\abs{x_a}{\letv{x_w}{\app{v_2}{x_a}&}{\return{(\indelimm{x_w}{H_1 \seq \indcase{op}{x_1}{x_2}{v_3} \seq H_2})}}}} \\
\app{(\abs{x}{c})}{v} &\stepstoc \sbst{c}{x}{v} \\
\letv{x}{\return{v}}{c} &\stepstoc \sbst{c}{x}{v} \\
\letv{x_2}{\letv{x_1}{\perform{\loc}{op}{v}}{c_1}}{c_2} &\stepstoc \letv{x_1}{\perform{\loc}{op}{v}}{\letv{x_2}{c_1}{c_2}} \\
\perform{(\outl{v_1})}{op}{v_2} &\stepstoc \perform{v_1}{\mathrm{left}(op)}{v_2} \\
\perform{(\outr{v_1})}{op}{v_2} &\stepstoc \perform{v_1}{\mathrm{right}(op)}{v_2} \\
\reify{\loc}{\return{(\gbox{v})}} &\stepstoc \return{\ret{v}} \\
\reify{\loc_1}{\perform{\loc_1}{op}{v}} &\stepstoc \return{\indintro{op}{v}{\abs{(\gbox{x})}{\return{\ret{x}}}}} \\
\reify{\loc_1}{\letv{x}{\perform{\loc_1}{op}{v}}{c}} &\stepstoc \return{\indintro{op}{v}{\abs{x}{\reify{\loc_1}{c}}}} \\
\reify{\loc_1}{\perform{\loc_2}{op}{v}} &\stepstoc \letv{(\gbox{x})}{\perform{\loc_2}{op}{v}}{\return{\ret{x}}} \qquad (\loc_1 \neq \loc_2) \\
\reify{\loc_1}{\letv{x}{\perform{\loc_2}{op}{v}}{c}} &\stepstoc \letv{x}{\perform{\loc_2}{op}{v}}{\reify{\loc_1}{c}} \qquad (\loc_1 \neq \loc_2)
\end{align*}
\caption{Reduction relations}
\label{fig:reduction}
\end{figure}
\begin{figure}
\begin{align*}
  \mathrm{neutral\ normal\ values}, \eta :=\;& x \mid \fst{\eta} \mid \snd{\eta} \mid \\
                                   &\casem{\eta}{x}{\omega}{x}{\omega} \mid \\
                                   &\gunbox{x}{\eta}{\omega} \mid \grun{\mu} \mid \\
                                   &\indelimm{\eta}
                                     {\indretcase{x}{\omega} \seq
                                     \indcase{op}{x}{x}{\omega} \seq \ldots \seq
                                     \indcase{op}{x}{x}{\omega}} \\
  \mathrm{normal\ values}, \omega :=\;& \eta \mid (\omega, \omega)
                           \mid () \mid \inl{\omega} \mid \inr{\omega} \mid \abs{x}{\mu} \mid \\
                           &\indintro{op}{\omega}{\omega} \mid \ret{\omega} \mid \\
                           &\gbox{\omega} \mid \loc \mid \outl{\omega} \mid \outr{\omega} \\
  \mathrm{neutral\ normal\ computations}, \nu :=\;& \app{\eta}{\omega}
                                            \mid \reify{\loc}{\nu} \mid \perform{\eta}{op}{\omega}
                                            \mid \letv{x}{\nu}{\mu} \\
  \mathrm{normal\ computations}, \mu :=\;& \nu \mid \return{\omega} \mid \perform{\loc}{op}{\omega}
                                              \mid \letv{x}{\perform{\loc}{op}{\omega}}{\mu}
\end{align*}
\caption{Normal forms}
\label{fig:normal}
\end{figure}

Note that the reduction rules are for strong reduction, i.e. there is a
congruence rule for abstraction:
\begin{mathpar}
\inferrule{c_1 \stepstoc c_2}
{\abs{x}{c_1} \stepstov \abs{x}{c_2}}
\end{mathpar}

We prove subject reduction for this semantics in
Appendix~\ref{apdx:subject-reduction} by induction on the reduction
relations.
\begin{lemma}
\label{lemma:subject-reduction}
  \begin{align*}
    &\Gamma \turnv v_1 \types \tau
      \quad \wedge \quad v_1 \stepstov v_2
      \quad \implies \quad \Gamma \turnv v_2 \types \tau \\
    &\Gamma \turnc c_1 \types \tau
      \quad \wedge \quad c_1 \stepstoc c_2
      \quad \implies \quad \Gamma \turnc c_2 \types \tau
  \end{align*}
\end{lemma}

These reduction relations are confluent:
\begin{lemma}
\label{lemma:church-rosser}
  \begin{align*}
    &v_1 \reducestov v_2
      \quad \wedge \quad v_1 \reducestov v_3
      \quad \implies \quad \exists\, v_4. \quad v_2 \reducestov v_4
      \quad \wedge \quad v_3 \reducestov v_4 \\
    &c_1 \reducestoc c_2
      \quad \wedge \quad c_1 \reducestoc c_3
      \quad \implies \quad \exists\, c_4. \quad c_2 \reducestoc c_4
      \quad \wedge \quad c_3 \reducestoc v_4 \\
  \end{align*}
\end{lemma}
which we prove using the standard approach via an equivalent notion of
parallel reduction Appendix~\ref{apdx:church-rosser}.

They are also strongly normalizing. We call the sets of values and
computations for which there exist no infinite chains
of reductions $\snv$ and $\snc$ respectively.
\begin{lemma}
\label{lemma:strong-normalization}
\begin{align*}
    &\Gamma \turnv v \types \tau
      \quad \implies \quad v \in \snv \\
    &\Gamma \turnc c \types \tau
      \quad \implies \quad c \in \snc \\
\end{align*}
\end{lemma}
We prove this using logical relations in
Appendix~\ref{apdx:strong-normalization}.

\subsection{Embedding the Simply-Typed Lambda Calculus}

Using the global modality and $\grun{c}$ construct we can embed the
Simply Typed Lambda Calculus (STLC) into this language as values. This
demonstrates that the language is a proper extension of STLC. We show this
embedding and prove it correct in Appendix~\ref{apdx:embedding}.

\section{The Ambient Monad model}
\label{sec:semantics}

The design of effect reflection falls out naturally from a novel
semantic model of algebraic effects. The key idea is to index the
meaning of types and terms by an ambient algebraic monad of effects:
rather than have the semantics of each type be a set of elements, the
semantics of each type is a mapping from the ambient monad to a set of
elements. We give a denotational semantics to the language from
Section~\ref{sec:language} using this model.

\subsection{Algebraic Theories and Free Algebras}

We need the monads we use for the ambient monad to have a well-defined
notion of sum. This is a requirement for composable effect handlers:
within the effect handler the ambient monad will be the sum of the
handled effect and the ambient monad outside of the hander. That rules
out allowing all monads as the ambient monad, since the sum of two
arbitrary monads may not exist.

Algebraic effects ensure that such a sum always exists by only
supporting monads that are the free algebra of some algebraic theory. We
use the same approach for the Ambient Monad model.

We recall the definitions of algebraic signature, terms of a signature,
algebraic theory and models of a theory.
\begin{definition}[Algebraic signature]
  An \emph{algebraic signature} is a set of operations with associated
  arities.
\end{definition}

\begin{example}
  The algebraic signature for boolean state $\Sigma_{bs}$ is:
  \begin{equation*}
    \{ (\mathrm{get}, 2); (\mathrm{set\_true}, 1); (\mathrm{set\_false}, 1) \}
  \end{equation*}
\end{example}

\begin{definition}[Term]
  A \emph{term} $t$ of some algebraic signature $\Sigma$ is either a
  free variable $x$ or $op(t_1, \ldots, t_k)$ where $op$ is an operation
  of $\Sigma$ with arity $k$ and each $t_i$ is itself a term.
\end{definition}

\begin{definition}[Algebraic theory]
  An \emph{algebraic theory} consists of an algebraic signature $\Sigma$ and a set of
  equations on terms of $\Sigma$.
\end{definition}

\begin{example}
  The algebraic theory for boolean state $L_{bs}$ adds the following
  equations to the signature of boolean state $\Sigma_{bs}$:
  \begin{align*}
    &\mathrm{set\_true}(\mathrm{get}(x, y)) = \mathrm{set\_true}(x)
    &\mathrm{set\_false}(\mathrm{get}(x, y)) = \mathrm{set\_false}(y) \\
    &\mathrm{set\_true}(\mathrm{set\_true}(x)) = \mathrm{set\_true}(x)
    &\mathrm{set\_true}(\mathrm{set\_false}(x)) = \mathrm{set\_false}(x) \\
    &\mathrm{set\_false}(\mathrm{set\_true}(x)) = \mathrm{set\_true}(x)
    &\mathrm{set\_false}(\mathrm{set\_false}(x)) = \mathrm{set\_false}(x) \\
    &\mathrm{get}(\mathrm{set\_true}(x), \mathrm{set\_false}(x))) = x
  \end{align*}
\end{example}

\begin{definition}[Model]
  A \emph{model} $M$ of some algebraic theory $L$ consists of a carrier
  set $S$ and for each operation $op$ in the signature of $L$ a function
  $op_M : (\Pi_k S) \rightarrow S$ where $k$ is the arity of $op$, such
  that the equations of $L$ are respected when interpreting the terms
  using these functions for the operations. Models of $L$ are also
  called algebras of $L$.
\end{definition}

\begin{example}
  We can model the theory of boolean state $L_{bs}$ using the carrier
  set $2 \rightarrow 2$ and interpreting the operations with:
  \begin{align*}
    &\mathrm{get}(x, y) = \lambda s.\, \mif s \mthen x(s) \melse y(s) \\
    &\mathrm{set\_true}(x) = \lambda s.\, x(\mathrm{true}) \\
    &\mathrm{set\_false}(x) = \lambda s.\, x(\mathrm{false})
  \end{align*}
\end{example}

Arities in algebraic signatures are often considered as natural numbers,
but they can equivalently be sets, with families of terms indexed by
that set used as the arguments of that operator in a term. It is also
convenient to allow families of operations parameterized by some
set. This gives us signatures of the form:
\begin{equation*}
\{ op_1 \types P_1 \rightarrowtail A_1; \ldots; op_n
\types P_n \rightarrowtail A_n \}
\end{equation*}
where $op \types P \rightarrowtail A$ indicates an operation $op$
parameterized by $P$ and with arity $A$.

\begin{example}
  The algebraic signature for boolean state $\Sigma_{bs}$ can be written
  as:
  \begin{equation*}
    \{ \mathrm{get} \types 1 \rightarrowtail 2; \mathrm{set} \types 2 \rightarrowtail 1 \}
  \end{equation*}
\end{example}

Given an algebraic theory $L$ and a set $A$ we can construct
the free model, or \emph{free algebra}, $T_L(A)$. The values $T_L(A)$
are terms built using the elements of $A$ and the operations of $L$
quotiented by the equations in $L$. $T_L$ is a monad and $T$ is one half
of an equivalence between algebraic theories and algebraic monads.

\begin{example}
  The free algebras of the signature of boolean state $T_{\Sigma_{bs}}$
  are equivalent to the type:
  \begin{lstlisting}[style=oxcaml]
  type 'a t =
    | Return of 'a
    | Get of unit * (bool -> 'a t)
    | Set of bool * (unit -> 'a t)
  \end{lstlisting}
  The free algebras of the theory of boolean state $T_{L_{bs}}$ are
  equivalent to the above type quotiented by the equations of $L_{bs}$.
\end{example}

For an algebraic signature $\Sigma$ the free algebra $T_{\Sigma}$ is the
free monad of an endofunctor $F_{\Sigma}$. The values of $F_{\Sigma}(A)$
are of the form $op(a_1, \ldots, a_k)$ where $op$ is an operation of
$\Sigma$ with arity $k$ and each $a_i$ is an element of $A$.
\begin{example}
  $F_{\Sigma_{bs}}$ is equivalent to the type:
  \begin{lstlisting}[style=oxcaml]
  type 'a t =
    | Get of unit * (bool -> 'a)
    | Set of bool * (unit -> 'a)
  \end{lstlisting}
\end{example}

\subsection{Lawvere theories}

\emph{Lawvere theories} are a categorical representation of algebraic
theories that doesn't depend on the particular choice of operations and
equations. We will use them as our concrete representation of the
ambient monad.

\begin{definition}
  A \emph{Lawvere theory} is a category $L$ with finite products in
  which every object is isomorphic to some finite cartesian product
  $X^n = X \times \ldots \times X$ of a distinguished object $X$.
\end{definition}
The objects of a Lawvere theory correspond to the different possible
arities and the morphisms $X^n \rightarrow X$ correspond to terms
of $L$ with $n$ free variables quotiented by the equations in $L$. A
model of a Lawvere theory $L$ is a product-preserving functor
$L \rightarrow \setc$.

We will mostly describe Lawvere theories and constructions on them in
terms of their presentations as algebraic theories, so an understanding
the details of Lawvere theories is not required for understanding the
bulk of this paper

Lawvere theories form a category $\lawfc$ whose morphisms are
equivalent to maps from the operations of one theory to the operations
of another that preserve the equations of the first theory according to
the equations of the second.

\subsection{Bounded Lawvere theories}
\label{sec:small}

Traditional algebraic signatures are restricted to the case of a finite
set of operations each with a finite arity, but programming languages
require more generality. In terms of effect reflection, this would
amonut to demanding, for any operation
$op \types \tau_p \rightarrowtail \tau_a$, that $\tau_p$ and $\tau_a$ be
finite types. This would allow for boolean state, but would not support
$\nat$ state.

In the algebraic effects literature the most common choice is to use
countable algebraic signatures. However, this is still too restritive,
supporting $\nat$ state but not supporting $\nat \rightarrow \nat$
state.

We want to allow arbitrary types for the parameters and arities of our
operations. However, we also wish our category of Lawvere theories to be
small. To achieve this we will parameterize our model and semantics by a
regular cardinal $\kappa$, and use algebraic theories whose set of
operations and arites are bound by $\kappa$.

We can extend the definition of Lawvere theory to support algebraic
theories bounded by $\kappa$:
\begin{definition}
  A \emph{$\kappa$-bounded Lawvere theory} is a small category whose set
  of objects has cardinality $\lambda < \kappa$ and whose set of
  morphims also has cardinality strictly less than $\kappa$, which has
  all products smaller than $\lambda$, and in which every object is
  isomorphic to some product $\Pi_{\alpha < \lambda} X$ of a
  distinguished object $X$.
\end{definition}
The models of a $\kappa$-bounded Lawvere theory $L$ are the
product-preserving functors from $L$ to $\setc$.

Morphisms between $\kappa$-bounded Lawvere theories are
product-preserving functors that preserve the distinguished object
$X$. Such functors correspond to homomorphisms of algebraic theories:
maps from the operations of one theory to the operations of another that
preserve the equations of the first theory according to the equations of
the second.

$\kappa$-bounded Lawvere theories and their morphisms form a locally
small category $\lawc$. This category has coproducts $L + L^{\prime}$
which correspond to the algebraic theory made by unioning the operations
and equations of $L$ and $L^{\prime}$. It also has an initial object
$\initial$, which is the algebraic theory with no operations and
corresponds to the identity monad. Note that $\lawfc$ is $\lawn$.

\subsection{Presheaves}

Now that we have a concrete representation for potential ambient monads,
we can define our semantics in terms of sets indexed by such
monads. More specifically, we can use functors $\setc^{\lawc}$ as the
semantics of our types. By using functors rather than ordinary functions
we ensure terms and types at a particular ambient monad can always be
mapped naturally to a larger ambient monad. These functors have many
convenient properties because they are \emph{presheaves}.

For a category $\cat$, $\psh = \setc^{\catop}$ is the category of
presheaves over $\cat$. Its objects are contravariant functors from
$\cat$ to $\setc$ and it's morphisms are natural transformations between
such functors.

In these terms, we use
\begin{equation*}
  \pshlawcop = \setc^{\lawc}
\end{equation*}
as the basis of the Ambient Monad model. The $\op$ in the definition of
presheaves and the $\op$ due to our dependence being contravariant
cancel out, but it is still profitable to think in terms of presheaves
as presheaves come with a lot of useful technology.

The Ambient Monad model uses objects from $\pshlawcop$ to represent
types and contexts, and morphisms in $\pshlawcop$ to represent
values.
\begin{equation*}
\sem{\tau} \mathrel{:} \pshlawcop \qquad
\sem{\Gamma} \mathrel{:} \pshlawcop \qquad
\sem{\Gamma \turnv v \types \tau} \mathrel{:} \homset{\pshlawcop}{\sem{\Gamma}}{\sem{\tau}}
\end{equation*}

A category of presheaves over a small category, such as $\lawc$, is a
topos, which gives it many useful properties. As a topos, $\pshlawcop$
has products and coproducts, which are defined pointwise. We will use
these as the semantics of product types and sum types.
\begin{equation*}
\sem{1} = 1 \qquad
\sem{\tau_1 \times \tau_2} = \sem{\tau_1} \times \sem{\tau_2} \qquad
\sem{\tau_1 + \tau_2} = \sem{\tau_1} + \sem{\tau_2} \qquad
\end{equation*}
The semantics of the corresponding introduction and elimination forms
follows the usual categorical semantics of such types.

We also use products for the semantics of contexts:
\begin{equation*}
\sem{\cempty} = 1 \qquad
\sem{\lbind{\Gamma}{x}{\tau}} = \sem{\Gamma} \times \sem{\tau}
\end{equation*}

\subsection{Representing computations: The free algebras monad}

Our aim is to have a semantics where the meaning of types is indexed by
the ambient monad -- or more precisely by a representation of the
algebraic theory whose free algebras are the ambient monad. In
particular, we would like the interpretation of computation types to be
applications of the ambient monad.

Given a $\kappa$-bounded Lawvere theory $L$, the underlying sets of its free
algebras form a monad $T_L$ on $\setc$. In addition to the unit of the
monad, we can construct elements of $T_L$ using the operations of
$L$. For an operation $op \types P \rightarrowtail A$ of a Lawvere
theory $L$ and some set $S$, there is a function
$op_{T_L} \types (P \times (A \rightarrow T_L(S))) \rightarrow T_L(S)$
that constructs terms in the free algebra using $op$.

$T_{\_}$ is functorial and we can use it to construct a monad $\mon$ on
$\pshlawcop$ such that:
\begin{equation*}
\mon(P)(L) = T_L(P(L))
\end{equation*}
We call $\mon$ the \emph{free algebras monad}, and it is the
representation of computations in the Ambient Monad model.
\begin{align*}
  &\sem{\Gamma \turnc v \types \tau} \mathrel{:}
    \homset{\pshlawcop}{\sem{\Gamma}}{\mon(\sem{\tau})} \\
  &\sem{\Gamma \turnc \return{v} \types \tau} =
    \unit{\mon}_{\sem{\tau}} \circ \sem{\Gamma \turnv v \types \tau} \\
  &\sem{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2} =
    \mult{\mon}_{\sem{\tau_2}}
    \circ \mon(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2})
    \circ \strength{\mon}_{\sem{\Gamma}, \sem{\tau_1}}
    \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnc c_1 \types \tau_1} \rangle
\end{align*}
where $\unit{\mon}$, $\mult{\mon}$ and $\strength{\mon}$ are the unit,
multiplication and strength respectively of $\mon$.

As a topos, $\pshlawcop$ also has exponentials, which combine with the
free algebras monad to give us our semantics for function types.
\begin{equation*}
  \sem{\tau_1 \arr \tau_2} = \arrow{\sem{\tau_1}}{\mon(\sem{\tau_2})}
\end{equation*}

\subsection{The global modality}

For any category $\cat$, if $\cat$ has a terminal object $\terminal$, we
can define the global modality $\glob \types [\psh, \psh]$ such that:
$\glob(P)(L) = P(\terminal)$. In the case of $\pshlawcop$ this becomes:
\begin{equation*}
\glob(P)(L) = P(\initial)
\end{equation*}
$\glob$ forms a comonad on $\pshlawcop$. $\glob$ also distributes over
products: $\glob (P \times Q) = (\glob P) \times (\glob Q)$

Within the abient monad model, this amounts to restricting a type to
only the elements which do not depend on the ambient monad. This is very
similar to the purity comonad of Choudhury and
Krishnaswami~\cite{choudhury2020recovering}. In particular,
\begin{equation*}
  \glob \mon(P) \quad = \quad T_{\initial} \circ \glob P
  \quad \cong \quad \glob P
\end{equation*}
as elements of the free algebras of the initial algebraic theory are
just elements of their carrier sets.

We use the $\glob$ comonad as our semantics for $\glob$ types and
$\glob$ bindings.
\begin{align*}
&\sem{\glob \tau} = \glob \sem{\tau} \\
&\sem{\gbind{\Gamma}{x}{\tau}} = \sem{\Gamma} \times \glob \sem{\tau} \\
&\sem{\gbind{\Gamma}{x}{\tau} \turnv x \types \tau} = \counit{\glob}_{\sem{\tau}} \circ \snd \\
&\sem{\Gamma \turnv \gbox{v} \types \glob \tau} =
  \glob \sem{\cdiv{\Gamma} \turnv v \types \tau}
  \circ \comult{\glob}_{\sem{\cdiv{\Gamma}}} \circ \bdiv{\Gamma} \\
&\sem{\Gamma \turnv \gunbox{x}{v_1}{v_2} \types \tau_2} =
  \sem{\gbind{\Gamma}{x}{\tau_1} \turnv v_2 \types \tau_2}
  \circ \langle id_{\sem{\Gamma}},
                \sem{\Gamma \turnv v_1 \types \glob \tau_1} \rangle \\
&\sem{\Gamma \turnv \grun{c} \types \glob \tau} =
  \ununit_{\sem{\tau}}
  \circ \glob \sem{\cdiv{\Gamma} \turnc c \types \tau}
  \circ \comult{\glob}_{\sem{\cdiv{\Gamma}}} \circ \bdiv{\Gamma}\\
\end{align*}
where $\counit{\glob}$ and $\comult{\glob}$ are the counit and
comultiplication of $\glob$,
$\bdiv{\Gamma} \types \Gamma \rightarrow \cdiv{\Gamma}$ is the morphism
that projects the global bindings from $\Gamma$, and $\ununit_P$ is the
isomorphism from $\glob \mon(P)$ to $\glob P$.

\subsection{Reflecting operations: The Yoneda embedding}

We want to be able to reflect operations into the ambient monad. We can
use the $op_{T_L}$ functions to build morphisms into $\mon$, but we
require some way to represent operations of the ambient monad. We can do
that by using the \emph{Yoneda embedding}. The Yoneda embedding
$\yoneda{\_}$ is a full and faithful embedding from a category $\cat$
into its category of presheaves $\psh$. When treating $\cat$ as a form
of context, the elements of $y(\catobj)$ can be used to represent
variables of type $\catobj$.

$\lawcop$ embeds into $\pshlawcop$ via the Yoneda embedding: for each
$\kappa$-bounded Lawvere theory $L$ there is a presheaf $\yoneda{L}$ in
$\pshlawcop$ such that:
\begin{align*}
&\yoneda{L}(L^{\prime}) = \homset{\lawc}{L}{L^{\prime}} \\
&\yoneda{L}(l^{\prime}) = \lambda f.\; l^{\prime} \circ f
\end{align*}

For an operation $op \types P \rightarrowtail A$ of a Lawvere theory
$L$, we can build a morphism:
\begin{align*}
&op_{\mon} \types
  \homset{\pshlawcop}
         {\yoneda{L} \times \glob P}
         {\mon{\glob A}} \\
&{(op_{\mon})}_{L^\prime}
  = \lambda(y, p). (y(op))_{T_{L^\prime}}(p, \unit{T_{L^\prime}}_{A})
\end{align*}
which uses an element of $\yoneda{L}$ to map $op$ into the ambient monad
and then uses that to construct a term in the free algebras monad.

We use $\yoneda{L}$ as the semantics for our effect capabilities, and
$op_{\mon}$ as the semantics for reflection.
\begin{align*}
&\sem{\yoneda{\Sigma}} = \yoneda{\sem{\Sigma}} \\
&\sem{\Gamma \turnc \perform{v_1}{op}{v_2} \types \tau_a} = \\
&\quad op_{\mon}
       \circ \langle \sem{\Gamma \turnv v_1 \types \yoneda{\Sigma}},
                      \sem{\Gamma \turnv v_2 \types \tau_p} \rangle
\end{align*}
where $\sem{\Sigma}$ is the $\kappa$-bounded Lawvere theory corresponding to the
signature $\Sigma$. If $\Sigma$ is:
\begin{equation*}
  \{ \, op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1} \seq
  \ldots \seq
  op_n \types \tau_{p_n} \rightarrowtail \tau_{a_n} \, \}
\end{equation*}
then $\sem{\Sigma}$ is the $\kappa$-bounded Lawvere theory for:
\begin{equation*}
  \{ \, op_1 \types \sem{\tau_{p_1}}(\initial)
                    \rightarrowtail \sem{\tau_{a_1}}(\initial) \seq
  \ldots \seq
  op_n \types \sem{\tau_{p_n}}(\initial)
              \rightarrowtail \sem{\tau_{a_n}}(\initial) \, \}
\end{equation*}
Note that by applying the types to the initial theory $\initial$ we
restrict them to their global elements: those that do not depend on the
current ambient monad.

The Yoneda embedding preserves finite products, which for our case
means:
\begin{equation*}
\yoneda{L + L^{\prime}} \cong \yoneda{L} \times \yoneda{L^{\prime}}
\end{equation*}
which we use for the semantics of $\outlname$ and $\outrname$.

\subsection{Representing effectful computations: The Yoneda lemma}

We can represent computations in a specific algebraic monad using the
Yoneda lemma. The Yoneda lemma says that, for any $P \types \psh$ and
$\catobj \types \cat$, there is an isomorphism:
\begin{equation*}
  \homset{\psh}{\yoneda{\catobj}}{P} \cong P(\catobj)
\end{equation*}

Applying that to the composition of the free algebras monad and the
global modality we get:
\begin{equation*}
  \homset{\pshlawcop}{\yoneda{L}}{\mon{\glob \tau}} \cong \mon{\glob \tau}(L) = T_L(\tau(\initial))
\end{equation*}

In terms of our semantics, this means that computations of type
$\glob \tau$ with a single free variable of type $\yoneda{\Sigma}$ are
isomorphic to values of the free algebra of $\sem{\Sigma}$ at
$\sem{\tau}(\initial)$.

\subsection{Effect reification and inductive types}

The essence of algebraic effect handlers is taking a computation in
$T_{L + \Sigma}(S)$ the free algebras of the sum of an algebraic theory
$L$ and an algebraic signature $\Sigma$, splitting out and handling
$\Sigma$, leaving a computation in only the algebraic theory
$T_{L}(S^\prime)$.

The key to this process is the following
lemma~\cite{hyland2006combining}.
\begin{lemma}
\label{lemma:sums}
  For any $\kappa$-bounded Lawvere theory $L$ and algebraic signature $\Sigma$
  \begin{equation*}
  T_{L + \Sigma}(S) \cong \mu X.T_L(S + F_\Sigma(X))
  \end{equation*}
\end{lemma}
where $\mu X.G(X)$ is the initial algebra of some polynomial endofunctor
$G$. This lemma is crucial because it shows that computations using both
ambient effects $L$ and the signature $\Sigma$ can be reorganized as a
recursive structure that separates the two effect sources -- exactly
what effect handlers need.

Effect reification works by exposing $\mu X.T_L(S + F_\Sigma(X))$ to the
user via an inductive type. In addition to being a topos, $\pshlawcop$
has a Natural Number Object, which we prove in \ref{apdx:NNO}. Thus it
has initial algebras of polynomial endofunctors
\cite{moerdijk2000wellfounded}, i.e. W-types, so we can construct the
required initial algebra as a presheaf. Given an algebraic signature
$\Sigma$ and a presheaf $P$, we define the following family of initial
algebras:
\begin{align*}
&\ind{\Sigma} \types [\pshlawcop, \pshlawcop] \\
&\ind{\Sigma}(P) = \mu X. ((\glob P) + F_\Sigma \circ \mon{X})
\end{align*}
and observe the following variation on Lemma \ref{lemma:sums}.
\begin{lemma}
\label{lemma:sums2}
  For any $\kappa$-bounded Lawvere theory $L$ and algebraic signature $\Sigma$
  \begin{equation*}
  T_{L + \Sigma}(P(\initial)) \cong \mon{(\ind{\Sigma}(P))}(L)
  \end{equation*}
\end{lemma}

We use these initial algebras as the semantics of $\ind{\Sigma}$:
\begin{mathpar}
\sem{\ind{\Sigma}(\tau)} = \ind{\sem{\Sigma}}(\sem{\tau})
\end{mathpar}

Exponentials in categories of presheaves are defined such that:
\begin{equation*}
\arrow{P}{Q}(\catobj) \cong \homset{\psh}{\yoneda{\catobj} \times P}{Q}
\end{equation*}
Applying this to $\arrow{\yoneda{L}}{\mon{(\glob P)}}$ and then applying
the Yoneda Lemma we get:
\begin{align*}
\arrow{\yoneda{L}}{\mon{(\glob P)}}(L^{\prime})
  &\cong \homset{\pshlawcop}{\yoneda{L^{\prime}} \times \yoneda{L}}{\mon{(\glob P)}} \\
  &\cong \homset{\pshlawcop}{\yoneda{L^{\prime} + L}}{\mon{(\glob P)}} \\
  &\cong \mon{(\glob P)}(L^{\prime} + L) \\
  &\cong T_{(L^{\prime} + L)}(P(\initial))
\end{align*}

In the case where $L$ is a signature $\Sigma$, we can further apply
Lemma \ref{lemma:sums2} and, observing that all the steps in this proof
were natural in $L^{\prime}$, produce this isomorphism:
\begin{equation*}
  \arrow{\yoneda{\Sigma}}{\mon{(\glob P)}} \cong \mon{(\ind{\Sigma}(P))}
\end{equation*}
which we call $\reifyc{\Sigma}{}$.

$\reifyc{\Sigma}{}$ gives us our semantics for reification.
\begin{align*}
&\sem{\Gamma \turnc \reify{\loc}{c} \types \ind{\Sigma}(\glob \tau)} =
  \reifyc{\Sigma}{\tau} \circ \currymap(\sem{\lbind{\Gamma}{\loc}{\Sigma} \turnc c \types \glob \tau}) \\
\end{align*}

\subsection{Properties of the denotational semantics}

The denotational semantics is sound:
\begin{lemma}
\label{lemma:soundness}
  \begin{equation*}
    \Gamma \turnv v_1 \types \tau
    \wedge v_1 \stepstov v_2
    \implies \sem{\Gamma \turnv v_1 \types \tau}
           = \sem{\Gamma \turnv v_2 \types \tau}
  \end{equation*}
\end{lemma}
and adequate for contextual equality $\ctxeq$:
\begin{lemma}
\label{lemma:adequacy}
  \begin{equation*}
    \sem{\turnv v_1 \types \tau} = \sem{\turnv v_2 \types \tau}
    \implies v_1 \ctxeq v_2
  \end{equation*}
\end{lemma}

We present the full semantics, precisely define contextual equality, and
prove these properties in Appendix~\ref{apdx:denotational}.

\section{Implementation in OxCaml}
\label{sec:practice}

We provide effect reflection as a simple library API in OxCaml based on
OxCaml's support for locality (see Section~\ref{sec:locality}). In this
section we describe this API and its implementation.

\subsection{Effect reflection interface}
\label{sec:interface}

The types of our the effect reflection interface are shown in
Fig.~\ref{fig:ox-types}.
\begin{figure}
\begin{lstlisting}[style=oxcaml]
  module type Sig = sig type 'a t end

  module Continuation(E : Sig) : sig
    type ('a, 'b) t
  end

  module Term (E : Sig) : sig
    type 'a t =
      | Return : 'a @@ global -> 'a t
      | Perform : 'r E.t @@ global * ('r, 'a) Continuation(E).t -> 'a t
  end

  module Handler(E : Sig) : sig
    type t
  end
\end{lstlisting}
\caption{Effect reflection types}
\label{fig:ox-types}
\end{figure}

These types are higher-kinded: they are parameterized by by type
constructors representing the operations of an algebraic effect. In
OxCaml, such types must be built using functors. We use a simple module
type \lstinline[style=oxcaml]{Sig} to act as the parameter type for
these functors.

The \lstinline[style=oxcaml]{'a Term(E).t} type corresponds to
$\ind{\Sigma}(\tau)$ and represents terms in the free algebra of
\lstinline[style=oxcaml]{E}.

The \lstinline[style=oxcaml]{Handler(E).t} corresponds to
$\yoneda{\Sigma}$ and represents the capability to perform operations
from \lstinline[style=oxcaml]{E}.

\lstinline[style=oxcaml]{('a, 'b) Continuation(E).t} is equivalent to
\lstinline[style=oxcaml]{'a -> 'b Term(E).t} when global and equivalent
to \lstinline[style=oxcaml]{'a -> 'b Term(E).t @ local} when local. This
ensures that the global modality distributes over
\lstinline[style=oxcaml]{Term(E).t}, which is an important property of
$\ind{\Sigma}$. The need for this type is an artifact of OxCaml's
approach to locality: the global modality does not distribute over the
function arrow in OxCaml.

\begin{figure}
\begin{lstlisting}[style=oxcaml]
  module Reflection (E : Sig) : sig
    val reify :
      (Handler(E).t @ local -> 'a) @ local -> 'a Term(E).t @ local

    val reify_global : (Handler(E).t @ local -> 'a) -> 'a Term(E).t

    val perform : 'a E.t -> (Handler(E).t @ local -> 'a)

    val continue :
      ('a, 'b) Continuation(E).t @ local -> 'a -> 'b Term(E).t @ local

    val continue_global : ('a, 'b) Continuation(E).t -> 'a -> 'b Term(E).t
  end
\end{lstlisting}
\caption{Effect reflection interface}
\label{fig:ox-interface}
\end{figure}
The main piece of the API is a family of isomorphisms:
\begin{equation*}
  \text{\lstinline[style=oxcaml]{'a Term(E).t}} \qquad
  \cong \qquad \text{\lstinline[style=oxcaml]{Handler(E).t @ local -> 'a}}
\end{equation*}
which we expose as three functions: \lstinline[style=oxcaml]{reify},
\lstinline[style=oxcaml]{reify_global} and
\lstinline[style=oxcaml]{perform}. These are provided by the
\lstinline[style=oxcaml]{Reflection} functor
(Fig.~\ref{fig:ox-interface}), along with the functions for continuing a
continuation.

\lstinline[style=oxcaml]{reify} and
\lstinline[style=oxcaml]{reify_global} implement the left-to-right
direction of the isomorphism. We need two separate functions because
OxCaml currently lacks any form of locality polymorphism.

\lstinline[style=oxcaml]{perform} is equivalent to the right-to-left
direction of the isomorphism. It reflects individual operations as
opposed to full terms, which is a more convenient primitive in
practice. We can implement reflection of \lstinline[style=oxcaml]{local}
full terms as:
\begin{lstlisting}[style=oxcaml]
  let rec reflect t h =
    match t with
    | Return x -> x
    | Perform(op, k) -> reflect (continue k (perform op h)) h
\end{lstlisting}
with type:
\begin{lstlisting}[style=oxcaml]
  val reflect : 'a Term(E).t -> (Handler(E).t @ local -> 'a)
\end{lstlisting}
Reflection of global terms can be defined
similarly.

The final piece of the interface is the support for selection from
effect capabilities for sums of signatures (i.e. $\outlname$ and
$\outrname$). These are provided via \lstinline[style=oxcaml]{Sum}
and \lstinline[style=oxcaml]{Select} functions shown in Fig.~\ref{fig:ox-selection}.
\begin{figure}
\begin{lstlisting}[style=oxcaml]
  module Sum (L : Sig) (R : Sig) : sig
    type 'a t =
      | Left : 'a L.t -> 'a t
      | Right : 'a R.t -> 'a t
  end

  module Select(L : Sig) (R : Sig) : sig
    val outl : Handler(Sum(L)(R)).t @ local -> Handler(L).t @ local

    val outr : Handler(Sum(L)(R)).t @ local -> Handler(R).t @ local
  end
\end{lstlisting}
\caption{Effect selection}
\label{fig:ox-selection}
\end{figure}

\subsection{Effect reflection implementation}

We can implement the effect reflection interface directly in terms of
OCaml's untracked effect
handlers~\cite{Sivaramakrishnan2021retrofitting}. The entire
implementation is only 44 lines of code and is given in
Fig.\ref{fig:implementation}. The trick is to generate a fresh effect
constructor for each reification, pass that constructor around as part
of the \lstinline[style=oxcaml]{handler}, and then use it to perform
operations during reflection. Performing effects involves a single
indirect call and then an underlying perform operation, so its
performance is essentially the same as the performance of OCaml's
algebraic effects.

For various practical reasons, OCaml's algebraic effects are restricted
to only allow continuations to be resumed a single time. This rules out
interesting effects like backtracking search, but improves the ability
to reason about the interaction of effects and resources.

There is no fundamental reason that effect reflection should be
restricted to only one-shot continuations, so for the purposes of this
paper we ignore the restriction. However, for the same reasons as
OCaml, our actual implementation is restricted one-shot
continuations. Note that OxCaml's \lstinline[style=oxcaml]{once}
mode~\cite{lorenzen2024oxidizing} allows us to enforce this restriction
statically.

\subsection{Examples}

We include additional examples and discussion in
Appendix~\ref{apdx:examples}. Of particular note are the handler for
Generators, which demonstrates the need for
\lstinline[style=oxcaml]{reify_global}, and the handler for the
combination of Generators and Asynchronous I/O, which demonstrates the
need for \lstinline[style=oxcaml]{outl} and
\lstinline[style=oxcaml]{outr}.

\section{Related work}
\label{sec:related-work}

\subsection{Lexical and named effect handlers}

Effect reflection is closely related to \emph{lexical effect handlers}
in systems such as Effekt~\cite{brachthauser2020effects} or Biernacki et
al.~\cite{biernacki2019binders}. Lexical effect handlers have a value
that represents the effect handler, similar to $\yoneda{\Sigma}$ values,
but these values are second class and require their own special typing
rules. More recent work on Effekt~\cite{brachthauser2022effects}
addresses the second-class nature of these values, but requires further
extending the type system by tracking the set of capabilities used by
each value.

Effect reflection is also closely related to \emph{named effect
  handlers}~\cite{xie2022first}. Named effect handlers use higher-rank
polymorphism to emulate locality tracking in much the same way as the
\lstinline[style=haskell]{St} monad in Haskell. This still requires
using an effect system to track which named effects are used by a
computation, in order to ensure that those effects do not escape the
scope of their handler. Using higher-rank polymorphism also prevents
global type inference, requiring complex inference that extends beyond
classic Hindley-Milner, and demanding a higher annotation burden on the
programmer.

\subsection{Monadic reflection}

Effect reflection is a variation on monadic
reflection~\cite{filinski1999representing}. By restricting ourselves to
algebraic monads with no equations we remove the need to manually define
\emph{glue} functions that describe the interaction between the monad
being reified and the current ambient monad. By using locality and
$\yoneda{\Sigma}$ values we avoid the need to have an effect system.

\subsection{Semantics of algebraic effect handlers}

Hyland et al.~\cite{hyland2006combining} investigate combining algebraic
effects, including the key property of sums of algebraic theories (our
Lemma \ref{lemma:sums}) on which handlers are based.

The original work on algebraic effect
handlers~\cite{plotkin2009handlers,plotkin2013handling} gives a
semantics in terms of free algebras but the semantics work with a fixed
algebraic theory: handlers do not actually change the ambient monad. Our
Ambient Monad model naturally accommodates changing ambient monad.

Bauer and Pretnar\cite{bauer2015programming} give a denotational
semantics for Eff but it is untyped. The same
authors\cite{bauer2015programming} also give a denotational semantics
for Eff that is typed, but where effects are treated as extrinsic to the
semantics.

Forster et al.~\cite{forster2017expressive} give a set-theoretic
denotational semantics of a calculus with algebraic effect handlers. It
is typed and represents effects using their free algebras. However,
their calculus supports neither nesting effect handlers nor effect
polymorphism, which means the handlers described are not modular or
composable.

\subsection{Comonadic purity}

Our use of a comonadic global modality to track values which do not
depend on the ambient monad is essentially the same as the purity
comonad described by Choudhury and
Krishnaswami~\cite{choudhury2020recovering}. Their semantic description
of their comonad is very different -- based on a semantics of
capabilities -- but the typing rules, and the practical result, are the
same.

\subsection{Locality and presheaves}

The Ambient Monad model is closely related to presheaf semantics for
higher-order abstract syntax~\cite{hu2022category}. In both cases
presheaves are used to index the semantics of types and terms by some
notion of context, with the Yoneda embedding used to access this context
and the global modality used to control dependence on it.

\bibliography{references}
\bibliographystyle{plain}

\appendix
\section{Operational semantics}

In this section we prove subject reduction, confluence and strong
normalizatoin for our operational semantics. We start by proving some
general lemmas that we'll need in the various proofs. Note that we will
refer to a property that applies to both value and computation terms,
and which must be shown by mutual induction, as a single lemma.

\subsection{Context division lemmas}

We will need some simple lemmas describing the behaviour of the context
division operator $\cdiv{\Gamma}$. All of which follow directly by
induction on the context.
\begin{lemma}
\label{lemma:local-cdiv}
  \begin{equation*}
    \cdiv{(\ccons{\lbind{\Gamma_1}{x}{v}}{\Gamma_2})} = \cdiv{(\ccons{\Gamma_1}{\Gamma_2})}
  \end{equation*}
\end{lemma}
\begin{lemma}
\label{lemma:yoneda-cdiv}
  \begin{equation*}
    \cdiv{(\ccons{\lbind{\Gamma_1}{\loc}{\Sigma}}{\Gamma_2})} = \cdiv{(\ccons{\Gamma_1}{\Gamma_2})}
  \end{equation*}
\end{lemma}
\begin{lemma}
\label{lemma:global-cdiv}
  \begin{equation*}
    \cdiv{(\ccons{\gbind{\Gamma_1}{x}{v}}{\Gamma_2})} = \ccons{\gbind{\cdiv{\Gamma_1}}{x}{v}}{\cdiv{\Gamma_2}}
  \end{equation*}
\end{lemma}
\begin{lemma}
\label{lemma:cdiv-idempotent}
  \begin{equation*}
    \cdiv{\cdiv{\Gamma}} = \cdiv{\Gamma}
  \end{equation*}
\end{lemma}

\subsection{Weakening lemmas}

We need lemmas stating that we can weaken typing judgements with new
variable bindings of both $x$ and $\loc$ variables.
\begin{lemma}
\label{lemma:local-weakening}
  \begin{align*}
    &\ccons{\Gamma_1}{\Gamma_2} \turnv v \types \tau_1
    \quad \implies \quad \ccons{\lbind{\Gamma_1}{x}{\tau_2}}{\Gamma_2} \turnv v \types \tau_1 \\
    &\ccons{\Gamma_1}{\Gamma_2} \turnc c \types \tau_1
    \quad \implies \quad \ccons{\lbind{\Gamma_1}{x}{\tau_2}}{\Gamma_2} \turnc c \types \tau_1
  \end{align*}
\end{lemma}
\begin{lemma}
\label{lemma:yoneda-weakening}
  \begin{align*}
    &\ccons{\Gamma_1}{\Gamma_2} \turnv v \types \tau
    \quad \implies \quad \ccons{\lbind{\Gamma_1}{\loc}{\Sigma}}{\Gamma_2} \turnv v \types \tau \\
    &\ccons{\Gamma_1}{\Gamma_2} \turnc c \types \tau
    \quad \implies \quad \ccons{\lbind{\Gamma_1}{\loc}{\Sigma}}{\Gamma_2} \turnc c \types \tau
  \end{align*}
\end{lemma}
The proofs of these follow by mutual induction on the value and
computation typing rules. The only aspect that isn't straightforward is
that rules with premises featuring $\cdiv{\Gamma}$ rely on
\ref{lemma:local-cdiv} and \ref{lemma:yoneda-cdiv} to reuse those
premises with the weakened context.

We similarly need a lemma stating that we can weakening typing
judgements with new global variable bindings.
\begin{lemma}
\label{lemma:global-weakening}
  \begin{align*}
    &\ccons{\Gamma_1}{\Gamma_2} \turnv v \types \tau_1
    \quad \implies \quad \ccons{\gbind{\Gamma_1}{x}{\tau_2}}{\Gamma_2} \turnv v \types \tau_1 \\
    &\ccons{\Gamma_1}{\Gamma_2} \turnc c \types \tau_1
    \quad \implies \quad \ccons{\gbind{\Gamma_1}{x}{\tau_2}}{\Gamma_2} \turnc c \types \tau_1
  \end{align*}
\end{lemma}
This again follows by mutual induction on the value and computation
typing rules. Rules with premises featuring $\cdiv{\Gamma}$ rely on
\ref{lemma:global-cdiv} in order to make use of the inductive hypothesis
on those premises.

We also need a lemma stating that we can weaken from $\cdiv{\Gamma}$ to
$\Gamma$.
\begin{lemma}
\label{lemma:cdiv-weakening}
  \begin{align*}
    &\ccons{\cdiv{\Gamma_1}}{\Gamma_2} \turnv v \types \tau
    \quad \implies \quad \ccons{\Gamma_1}{\Gamma_2} \turnv v \types \tau \\
    &\ccons{\cdiv{\Gamma_1}}{\Gamma_2} \turnc c \types \tau
    \quad \implies \quad \ccons{\Gamma_1}{\Gamma_2} \turnc c \types \tau \\
  \end{align*}
\end{lemma}
This follows by induction on $\Gamma_1$ via the local weakening lemmas
(\ref{lemma:local-weakening} and \ref{lemma:yoneda-weakening}).

\subsection{Substitution lemmas}

We need a simple lemma stating that substituting unbound variables does
nothing, which follows directly from mutual induction on typing rules:
\begin{lemma}
\label{lemma:unbound-substitution}
  \begin{align*}
    &\Gamma \turnv v_1 \types \tau
    \quad \wedge \quad x \notin \Gamma
    \quad \implies \quad \sbst{v_1}{x}{v_2} = v_1 \\
    &\Gamma \turnc c \types \tau
    \quad \wedge \quad x \notin \Gamma
    \quad \implies \quad \sbst{c}{x}{v} = c
  \end{align*}
\end{lemma}

We need a lemma demonstrating that substitution for local bindings
preserves typing.
\begin{lemma}
\label{lemma:local-substitution}
  \begin{align*}
    &\lbind{\Gamma}{x}{\tau_1} \turnv v_1 \types \tau_2
    \quad \wedge \quad \Gamma \turnv v_2 \types \tau_1
    \quad \implies \quad \Gamma \turnv \sbst{v_1}{x}{v_2} \types \tau_2 \\
    &\lbind{\Gamma}{x}{\tau_1} \turnc c \types \tau_2
    \quad \wedge \quad \Gamma \turnv v \types \tau_1
    \quad \implies \quad \Gamma \turnv \sbst{c}{x}{v} \types \tau_2
  \end{align*}
\end{lemma}
This follows by mutual induction on the typing rules for $v_1$ and
$c$. Cases with premises that extend the context rely on weakening
(\ref{lemma:local-weakening} and \ref{lemma:global-weakening}) to apply
the inductive hypothesis on those premises. Cases with premises that use
$\cdiv{}$ rely on \ref{lemma:unbound-substitution} to reuse the original
premise.

Similarly, we need a lemma demonstrating that substitution for global bindings
preserves typing.
\begin{lemma}
\label{lemma:global-substitution}
  \begin{align*}
    &\gbind{\Gamma}{x}{\tau_1} \turnv v_1 \types \tau_2
    \quad \wedge \quad \cdiv{\Gamma} \turnv v_2 \types \tau_1
    \quad \implies \quad \Gamma \turnv \sbst{v_1}{x}{v_2} \types \tau_2 \\
    &\gbind{\Gamma}{x}{\tau_1} \turnc c \types \tau_2
    \quad \wedge \quad \cdiv{\Gamma} \turnv v \types \tau_1
    \quad \implies \quad \Gamma \turnc \sbst{c}{x}{v} \types \tau_2
  \end{align*}
\end{lemma}
This follows by mutual induction on the typing rules for $v_1$ and
$c$. Cases with premises that extend the context with global bindings
rely on weakening (\ref{lemma:global-weakening}) to apply the inductive
hypothesis on those premises. Cases with premises that extend the
context with local bindings can already apply the inductive hypothesis
without weakening. Cases with premises that use $\cdiv{}$ rely on the
idempotence of $\cdiv{}$ (\ref{lemma:cdiv-idempotent}) to apply the
inductive hypothesis. The variable case that actually substitutes $v_2$
uses Lemma \ref{lemma:cdiv-weakening} to weaken $v_2$ to the right
context.

When reasoning about reductions we will need a lemma stating that
substitutions commute:
\begin{lemma}
\label{lemma:substitutions-commute}
  \begin{align*}
    &\sbst{\sbst{v_3}{x_1}{v_1}}{x_2}{v_2} = \sbst{\sbst{v_3}{x_2}{v_2}}{x_1}{\sbst{v_1}{x_2}{v_2}} \\
    &\sbst{\sbst{c}{x_1}{v_1}}{x_2}{v_2} = \sbst{\sbst{c}{x_2}{v_2}}{x_1}{\sbst{v_1}{x_2}{v_2}}
  \end{align*}
\end{lemma}
which follows directly by mutual induction on $v_3$ and $c$.

\subsection{Subject reduction}
\label{apdx:subject-reduction}

We prove subject reduction (Lemma~\ref{lemma:subject-reduction}):
\begin{align*}
  &\Gamma \turnv v_1 \types \tau
  \quad \wedge \quad v_1 \stepstov v_2
  \quad \implies \quad \Gamma \turnv v_2 \types \tau \\
  &\Gamma \turnc c_1 \types \tau
  \quad \wedge \quad c_1 \stepstoc c_2
  \quad \implies \quad \Gamma \turnc c_2 \types \tau
\end{align*}
by mutual induction over the two reduction relations. The reduction
relations are the least relations closed under the rules in
Fig.~\ref{fig:reduction} combined with the obvious congruence rules.

\begin{proof}
As all the typing rules are entirely compositional, the cases for the
congruence rules all follow directly by induction. This leaves only the
rules from Fig.~\ref{fig:reduction}.

Note that the typing rules are syntax directed so there is always only a
single typing rule that we need to consider.

\begin{description}
\item[Products] For the product beta rules:
\begin{align*}
\fst{(v_1, v_2)} &\stepstov v_1 \\
\snd{(v_1, v_2)} &\stepstov v_2
\end{align*}
the cases follow directly by inversion on the typing rule.

\item[Sums] For the sum rules:
\begin{align*}
\casem{(\inl{v_1})}{x_1}{v_2}{x_2}{v_3} &\stepstov \sbst{v_2}{x_1}{v_1} \\
\casem{(\inr{v_1})}{x_1}{v_2}{x_2}{v_3} &\stepstov \sbst{v_3}{x_2}{v_1}
\end{align*}
the cases follow by inversion on the typing rules followed by case
splitting on the modality $m$ and applying the either the global
(\ref{lemma:global-substitution}) or local
(\ref{lemma:local-substitution}) substition lemma.

\item[Global modality] For the global modality rule:
\begin{equation*}
\gunbox{x}{\gbox{v_1}}{v_2} \stepstov \sbst{v_2}{x}{v_1}
\end{equation*}
the case follows by inversion on the typing rules followed by the global
substitution lemma (\ref{lemma:global-substitution}).

For the run rule:
\begin{equation*}
\grun{\return{v}} \stepstov v
\end{equation*}
the case follows by inversion on the typing rules along with weakening
via Lemma \ref{lemma:cdiv-weakening}.

\item[Inductive types]
For the $\indelimname$/$\retname$ rule:
\begin{equation*}
\indelimm{(\ret{v_s})}{\indretcase{x}{v_e} \seq \ldots} \stepstov \sbst{v_e}{x}{v_s}
\end{equation*}
the case follows by inversion on the typing rules followed by the global
substitution lemma (\ref{lemma:global-substitution}).

For the $\indelimname$/$op$ rule:
\begin{align*}
\indelimm{(\indintro{op}{v_1}{v_2})}{H_1 \seq \indcase{op}{x_1}{x_2}{v_3} \seq H_2} &\stepstov \\
  \sbst{\sbst{v_3}{x_1}{v_1}}{x_2}{\abs{x_a}{\letv{x_w}{\app{v_2}{x_a}&}{\return{(\indelimm{x_w}{H_1 \seq \indcase{op}{x_1}{x_2}{v_3} \seq H_2})}}}}
\end{align*}
the case follows by inversion on the typing rules followed by the global
substitution lemma (\ref{lemma:global-substitution}) for $x_1$ and the
local substition lemma (\ref{lemma:local-substitution}) for $x_2$.

\item[Functions]
The case for the function beta rule:
\begin{equation*}
\app{(\abs{x}{c})}{v} \stepstoc \sbst{c}{x}{v}
\end{equation*}
follows by inversion on the typing rules followed by the local
substitution lemma (\ref{lemma:local-substitution}).

\item[Bind]
The case for the bind/return rule:
\begin{equation*}
\letv{x}{\return{v}}{c} \stepstoc \sbst{c}{x}{v}
\end{equation*}
follows by inversion on the typing rules followed by the local
substitution lemma (\ref{lemma:local-substitution}).

The case for the bind commuting rule: 
\begin{equation*}
\letv{x_2}{\letv{x_1}{\perform{\loc}{op}{v}}{c_1}}{c_2} \stepstoc \letv{x_1}{\perform{\loc}{op}{v}}{\letv{x_2}{c_1}{c_2}}
\end{equation*}
follows by inversion on the typing rules followed by weakening $c_3$ by
the local binding of $x_1$.

\item[Reflection]

The cases for the $\outlname$ and $\outrname$ rules
\begin{align*}
\perform{(\outl{v_1})}{op}{v_2} &\stepstoc \perform{v_1}{\mathrm{left}(op)}{v_2} \\
\perform{(\outr{v_1})}{op}{v_2} &\stepstoc \perform{v_1}{\mathrm{right}(op)}{v_2}
\end{align*}
follow directly by inversion on the typing rules.

\item[Reification]

The cases for the $\reifyname$ rules:
\begin{align*}
\reify{\loc}{\return{(\gbox{v})}} &\stepstoc \return{\ret{v}} \\
\reify{\loc_1}{\perform{\loc_1}{op}{v}} &\stepstoc \return{\indintro{op}{v}{\abs{(\gbox{x})}{\return{\ret{x}}}}} \\
\reify{\loc_1}{\letv{x}{\perform{\loc_1}{op}{v}}{c}} &\stepstoc \return{\indintro{op}{v}{\abs{x}{\reify{\loc_1}{c}}}} \\
\reify{\loc_1}{\perform{\loc_2}{op}{v}} &\stepstoc \letv{(\gbox{x})}{\perform{\loc_2}{op}{v}}{\return{\ret{x}}} \qquad (\loc_1 \neq \loc_2) \\
\reify{\loc_1}{\letv{x}{\perform{\loc_2}{op}{v}}{c}} &\stepstoc \letv{x}{\perform{\loc_2}{op}{v}}{\reify{\loc_1}{c}} \qquad (\loc_1 \neq \loc_2)
\end{align*}
follow by inversion on the typing rules followed by \ref{lemma:yoneda-cdiv}.
\end{description}
\end{proof}

\subsection{Confluence}
\label{apdx:church-rosser}

We prove that the reduction relations are confluent:
\begin{align*}
  &v_1 \reducestov v_2
    \quad \wedge \quad v_1 \reducestov v_3
    \quad \implies \quad \exists\, v_4. \quad v_2 \reducestov v_4
    \quad \wedge \quad v_3 \reducestov v_4 \\
  &c_1 \reducestoc c_2
    \quad \wedge \quad c_1 \reducestoc c_3
    \quad \implies \quad \exists\, c_4. \quad c_2 \reducestoc c_4
    \quad \wedge \quad c_3 \reducestoc v_4 \\
\end{align*}

As is standard, we start by constructing notions of parallel reduction
$\parastepstov$ and $\parastepstoc$. The inference rules describing
parallel reduction are given in Fig.~\ref{fig:parallel-value-reduction}
and Fig.~\ref{fig:parallel-computation-reduction}. We omit the
confluence rules. We write $\parareducestov$ and $\parareducestoc$ for
the transitive closures of $\parastepstov$ and $\parastepstoc$
respectively.
\begin{figure}
\begin{mathpar}
\inferrule{v_1 \parastepstov v_1^{\prime}}
{\fst{(v_1, v_2)} \parastepstov v_1^{\prime}}
\and
\inferrule{v_2 \parastepstov v_2^{\prime}}
{\snd{(v_1, v_2)} \parastepstov v_2^{\prime}}
\and
\inferrule{v_1 \parastepstov v_1^{\prime} \\ v_2 \parastepstov v_2^{\prime}}
{\casem{(\inl{v_1})}{x_1}{v_2}{x_2}{v_3} \parastepstov \sbst{v_2^{\prime}}{x_1}{v_1^{\prime}}}
\and
\inferrule{v_1 \parastepstov v_1^{\prime} \\ v_3 \parastepstov v_3^{\prime}}
{\casem{(\inr{v_1})}{x_1}{v_2}{x_2}{v_3} \stepstov \sbst{v_3^{\prime}}{x_2}{v_1^{\prime}}}
\and
\inferrule{v_1 \parastepstov v_1^{\prime} \\ v_2 \parastepstov v_2^{\prime}}
{\gunbox{x}{\gbox{v_1}}{v_2} \stepstov \sbst{v_2^{\prime}}{x}{v_1^{\prime}}}
\and
\inferrule{v \parastepstov v^{\prime}}
{\grun{\return{v}} \stepstov v^{\prime}}
\and
\inferrule{v_1 \parastepstov v_1^{\prime} \\ v_2 \parastepstov v_2^{\prime}}
{\indelimm{(\ret{v_1})}{\indretcase{x}{v_2} \seq \ldots} \stepstov \sbst{v_2^{\prime}}{x}{v_1^{\prime}}}
\and
\inferrule{v_p \parastepstov v_p^{\prime} \\ v_k \parastepstov v_k^{\prime} \\ v_e \parastepstov v_e^{\prime} \\ \overline{v_j \parastepstov v_j^{\prime}}}
{\indelimm{\indintro{op_i}{v_p}{v_k}}{\indretcase{x_r}{v_e} \seq
                                   \overline{\indcase{op_j}{x_{p_j}}{x_{k_j}}{v_j}}} \\\\ \parastepstov \\\\
  \sbst{\sbst{v_i^{\prime}}{x_{p_i}}{v_p^{\prime}}}{x_{k_i}}{\abs{x_a}{\letv{x_w}{\app{v_k^{\prime}}{x_a}}{
     \return{(\indelimm{x_w}{\indretcase{x_r}{v_e^{\prime}} \seq
                \overline{\indcase{op_j}{x_{p_j}}{x_{k_j}}{v_j^{\prime}}}})}}}}}
\end{mathpar}
\caption{Parallel reduction relations (values)}
\label{fig:parallel-value-reduction}
\end{figure}

\begin{figure}
\begin{mathpar}
\inferrule{c \parastepstoc c^{\prime} \\ v \parastepstov v^{\prime}}
{\app{(\abs{x}{c})}{v} \parastepstoc \sbst{c^{\prime}}{x}{v^{\prime}}}
\and
\inferrule{v \parastepstov v^{\prime} \\ c \parastepstoc c^{\prime}}
{\letv{x}{\return{v}}{c} \parastepstoc \sbst{c}{x}{v}}
\and
\inferrule{v \parastepstoc v^{\prime} \\ c_1 \parastepstoc c_1^{\prime} \\ c_2 \parastepstoc c_2^{\prime}}
{\letv{x_2}{\letv{x_1}{\perform{\loc}{op}{v}}{c_1}}{c_2} \parastepstoc \letv{x_1}{\perform{\loc}{op}{v^{\prime}}}{\letv{x_2}{c_1^{\prime}}{c_2^{\prime}}}}
\and
\inferrule{v_1 \parastepstov v_1^{\prime} \\ v_2 \parastepstov v_2^{\prime}}
{\perform{(\outl{v_1})}{op}{v_2} \parastepstoc \perform{v_1^{\prime}}{\mathrm{left}(op)}{v_2^{\prime}}}
\and
\inferrule{v_1 \parastepstov v_1^{\prime} \\ v_2 \parastepstov v_2^{\prime}}
{\perform{(\outr{v_1})}{op}{v_2} \parastepstoc \perform{v_1^{\prime}}{\mathrm{right}(op)}{v_2^{\prime}}}
\and
\inferrule{v \parastepstov v^{\prime}}
{\reify{\loc}{\return{(\gbox{v})}} \parastepstoc \return{\ret{v^{\prime}}}}
\and
\inferrule{v \parastepstov v^{\prime}}
{\reify{\loc_1}{\perform{\loc_1}{op}{v}} \parastepstoc \return{\indintro{op}{v^{\prime}}{\abs{(\gbox{x})}{\return{\ret{x}}}}}}
\and
\inferrule{v \parastepstov v^{\prime} \\ c \parastepstoc c^{\prime}}
{\reify{\loc_1}{\letv{x}{\perform{\loc_1}{op}{v}}{c}} \parastepstoc \return{\indintro{op}{v^{\prime}}{\abs{x}{\reify{\loc_1}{c^{\prime}}}}}}
\and
\inferrule{v \parastepstov v^{\prime}}
{\reify{\loc_1}{\perform{\loc_2}{op}{v}} \parastepstoc \letv{(\gbox{x})}{\perform{\loc_2}{op}{v^{\prime}}}{\return{\ret{x}}} \qquad (\loc_1 \neq \loc_2)}
\and
\inferrule{v \parastepstov v^{\prime} \\ c \parastepstoc c^{\prime}}
{\reify{\loc_1}{\letv{x}{\perform{\loc_2}{op}{v}}{c}} \parastepstoc \letv{x}{\perform{\loc_2}{op}{v^{\prime}}}{\reify{\loc_1}{c^{\prime}}} \qquad (\loc_1 \neq \loc_2)}
\end{mathpar}
\caption{Parallel reduction relations (computations)}
\label{fig:parallel-computation-reduction}
\end{figure}

Parallel reductions are reflexive:
\begin{lemma}
\label{lemma:parallel-reduction-reflexive}
  \begin{equation*}
    v \parastepstov v \qquad
    c \parastepstoc c
  \end{equation*}
\end{lemma}
which follows by induction on $v$ and $c$ via the confluence rules.

A single step of ordinary reduction is a valid parallel reduction:
\begin{lemma}
\label{lemma:parallel-reduction-of-reduction}
  \begin{align*}
    &v_1 \stepstov v_2
    \quad \implies \quad v_1 \parastepstov v_2 \\
    &c_1 \stepstoc c_2
    \quad \implies \quad c_1 \parastepstoc c_2
  \end{align*}
\end{lemma}
This follows by induction on the ordinary reductions. Each rule of the
ordinary reduction has a corresponding rule in the parallel reduction
that can be used, applying reflexivity of the parallel reduction
\ref{lemma:parallel-reduction-reflexive} on subterms as needed.

A single step of parallel reduction corresponds to a sequence of steps
in the ordinary reduction.
\begin{lemma}
\label{lemma:reductions-of-parallel-reduction}
  \begin{align*}
    &v_1 \parastepstov v_2
    \quad \implies \quad v_1 \reducestov v_2 \\
    &c_1 \parastepstoc c_2
    \quad \implies \quad c_1 \reducestoc c_2
  \end{align*}
\end{lemma}
This follows by induction on the parallel reductions. Each rule of the
parallel reduction relation has a corresponding rule of the ordinary
reduction relation. We use transitivity of $\reducestoname$ to first
apply confluence rules to reduce subterms before applying the actual
corresponding rule.

\subsubsection{Substitution}

The above lemmas mean that showing confluence of $\parareducestoname$ is
sufficient to show confluence of $\reducestoname$. To do that we first
need a lemma showing that substitution commutes with parallel reduction:
\begin{lemma}
\label{lemma:subst-of-parallel-reduction}
  \begin{align*}
  &v_1 \parastepstov v_3
    \quad \wedge \quad v_2 \parastepstov v_4
    \quad \implies \quad \sbst{v_1}{x}{v_2} \parastepstov \sbst{v_3}{x}{v_4} \\
  &c_1 \parastepstov c_2
    \quad \wedge \quad v_1 \parastepstov v_2
    \quad \implies \quad \sbst{c_1}{x}{v_1} \parastepstov \sbst{c_2}{x}{v_2} \\
  \end{align*}
\end{lemma}
\begin{proof}
We prove this by mutual induction on the parallel reduction relations on
$v_1$ and $c_1$.
\begin{description}
\item[Variables]
  If the variable is $x$ then this case follows directly from $v_2 \parastepstov v_4$,
  otherwise it follows directly from $v_1 \parastepstov v_3$.
\item[Other confluence rules] For confluence rules other than the
  variable rule we apply the same confluence rule and use the inductive
  hypothesis for any subcomponents.
\item[Reduction rules] For reduction rules we apply the same reduction
  rule, commute any substitutions produced by the reduction with the
  substitution of $x$ using \ref{lemma:substitutions-commute}, and then
  use the inductive hypothesis for any subcomponents.
\end{description}
\end{proof}

\subsubsection{The diamond property}

We can now show confluence of $\parareducestoname$ by proving the
stronger property that $\parastepstoname$ satisfies the diamond
property:
\begin{lemma}
\label{lemma:parallel-reduction-diamond}
\begin{align*}
  &v_a \parastepstov v_b
    \quad \wedge \quad v_a \parastepstov v_c
    \quad \implies \quad \exists\, v_d. \quad v_b \parastepstov v_d
    \quad \wedge \quad v_c \parastepstov v_d \\
  &c_a \parastepstoc c_b
    \quad \wedge \quad c_a \parastepstoc c_c
    \quad \implies \quad \exists\, c_d. \quad c_b \parastepstoc c_d
    \quad \wedge \quad c_c \parastepstoc c_d \\
\end{align*}
\end{lemma}
\begin{proof}
  We prove this by mutual induction on the parallel reductions to $v_b$
  and $c_b$, followed by inversion of the parallel reductions to $v_c$
  and $c_c$. As all congruence rules operate on different constructs,
  and no two reduction rule can apply to the same term, there are only
  two cases we must consider:
\begin{description}
\item[The same rule] If both reductions are the same rule then the
  induction hypotheses gives us a value and associated parallel
  reductions for each subcomponent. We use these values to build
  $v_d$/$c_d$ and apply the rule to build the required parallel
  reductions.
\item[A reduction rule and a confluence rule] Cases where
  one reduction is a reduction rule and the other is a confluence rule
  then all follow the same reasoning. We illustrate this reasoning by
  describing the case for the $\letvname/\returnname$ reduction rule.

  We have $v_a = \letv{x_l}{\return{v_{ar}}}{c_{as}}$,
  $v_b = \sbst{c_{bs}}{x_l}{v_{br}}$ and
  $v_c = \letv{x_l}{v_{cl}}{c_{cs}}$, and the rules in
  question are:
  \begin{mathpar}
    \inferrule{v_{ar} \parastepstov v_{br} \\ c_{as} \parastepstoc c_{bs}}
    {\letv{x_l}{\return{v_{ar}}}{c_{as}} \parastepstoc \sbst{c_{bs}}{x_l}{v_{br}}}
    \and
    \inferrule{\return{v_{ar}} \parastepstov v_{cl} \\ c_{as} \parastepstoc c_{cs}}
    {\letv{x_l}{\return{v_{ar}}}{c_{as}} \parastepstoc \letv{x}{v_{cl}}{c_{cs}}}
  \end{mathpar}
  All our reduction rules have a single subcomponent that is not a
  metavariable -- in this case it is $\return{v}$ -- and there are no
  reduction rules which could apply to that subcomponent. That means
  we know by inversion that $\return{v_{ar}} \parastepstov v_{cl}$ must be the confluence rule on $\returnname$,
  which gives us $v_{cl} = \return{v_{cr}}$, and the combined confluence rules amount to:
  \begin{mathpar}
    \inferrule{v_{ar} \parastepstov v_{cr} \\ c_{as} \parastepstoc c_{cs}}
    {\letv{x_l}{\return{v_{ar}}}{c_{as}} \parastepstoc \letv{x}{\return{v_{cr}}}{c_{cs}}}
  \end{mathpar}
  For some of the cases, e.g. the $\reifyname$/$\returnname$ rule, the
  subcomponent itself has a nested subcomponent that is not a
  metavariable and this inversion step must be repeated a second time.

  By the inductive hypothesis, we know that:
  \begin{align*}
  \exists\, v_{dr}. \,v_{br} \parastepstov v_{dr} \, \wedge \, v_{cr} \parastepstov v_{dr} \\
  \exists\, c_{ds}. \,c_{bs} \parastepstov c_{ds} \, \wedge \, v_{cs} \parastepstov v_{ds}
  \end{align*}
  We take $v_d$ to be $\sbst{c_{ds}}{x_l}{v_{dr}}$. We have
  $\sbst{c_{bs}}{x_l}{v_{br}} \parastepstov \sbst{c_{ds}}{x_l}{v_{dr}}$
  using \ref{lemma:subst-of-parallel-reduction}, and we have
  $\letv{x}{\return{v_{cr}}}{c_{cs}} \parastepstov \sbst{c_{ds}}{x_l}{v_{dr}}$
  by applying the reduction rule. In some cases there is no
  substitution and so the lemma is not needed, and in some other cases
  the result of the reduction is not a metavariable and so we must
  first apply the relevant confluence rule.
\end{description}
\end{proof}

\subsection{Strong normalization}
\label{apdx:strong-normalization}

\begin{align*}
    &\Gamma \turnv v \types \tau
      \quad \implies \quad v \in \snv \\
    &\Gamma \turnc c \types \tau
      \quad \implies \quad c \in \snc \\
\end{align*}

We prove strong normalization in the standard way for a simply typed
calculus via logical relations. In addition to being indexed by types
these relations will be parameterized by an \emph{effect context}: a
context restricted to bindings of the form $\loc \types \Sigma$.

\begin{equation*}
  \mathrm{effect\ contexts}, \Delta ::=\; \cempty \mid \lbind{\Gamma}{\loc}{\Sigma}
\end{equation*}

We write $\clocs{\Gamma}$ for the effect context obtained by removing
all the ordinary variable bindings from $\Gamma$ and keeping only the
location bindings.

As in Girard~\cite{girard1989proofs}, we will use the fact that strongly normalizable
terms must have a bound on the length of their possible reduction
sequences.
\begin{lemma}
\label{lemma:sn-bound}
For any strongly normalisable value $v$ (or computation $c$) there is a
number $\bound{v}$ (or $\bound{c}$) which bounds the length of every
normalization sequence beginning with $v$ (or $c$).
\end{lemma}
\begin{proof}
We can represent the possible normalization sequences as a tree, where the nodes
are reducts. Since there are a finite number of subterms of any value this tree
must be finitely branching. As the term is strongly normalizing it cannot contain
any infinite branches. Thus, by K\H{o}nig's lemma, the whole tree is finite and
there must be a $\bound{v}$ that bounds its height.
\end{proof}

We define a pair of mutually defined logical relations
$\redva{\Delta}{\tau}{\alpha}$ and $\redca{\Delta}{\tau}{\alpha}$
indexed by an effect context, a type and an ordinal. The ordinal
controls the rank of the well-founded trees corresponding to inductive
types and computations. The definitions are given in
Fig.~\ref{fig:reducibility-predicates}.

We write $\redv{\Delta}{\tau}$ for
$\mathrm{lim}_{\alpha} \redva{\Delta}{\tau}{\alpha}$ and
$\redc{\Delta}{\tau}$ for
$\mathrm{lim}_{\alpha} \redca{\Delta}{\tau}{\alpha}$. These limits are
well-defined because the predicates are monotonic in $\alpha$.

\begin{figure}
\begin{align*}
    \redva{\Delta}{1}{\alpha} \quad &= \quad
      \snv \\
    \redva{\Delta}{\tau_1 \times \tau_2}{\alpha} \quad &= \quad
       \{ \, v \; | \; \fst{v} \in \redv{\Delta}{\tau_1} \; \wedge \; \snd{v} \in \redv{\Delta}{\tau_2} \, \} \\
    \redva{\Delta}{\tau_1 + \tau_2}{\alpha} \quad &= \quad
       \{ \, v_1 \in \snv \; | \; \forall v_2. \; v_1 \reducestov \inl{v_2} \; \implies \; v_2 \in \redv{\Delta}{\tau_1} \\
         &\qquad \qquad \qquad \wedge \; \forall v_2. \; v_1 \reducestov \inr{v_2} \; \implies \; v_2 \in \redv{\Delta}{\tau_2} \, \} \\
    \redva{\Delta}{\tau_1 \arr \tau_2}{\alpha} \quad &= \quad
      \{ \, v_1 \; | \; \forall v_2 \in \redv{\Delta}{\tau_1}. \; \app{v_1}{v_2} \in \redc{\Delta}{\tau_2} \, \} \\
    \redva{\Delta}{\glob \tau}{\alpha} \quad &= \quad
      \{ \, v_1 \in \snv \; | \; \forall v_2. \; v_1 \reducestov \gbox{v_2} \; \implies \; v_2 \in \redv{\cempty}{\tau} \, \} \\
    \redva{\Delta}{\yoneda{\Sigma}}{\alpha} \quad &= \quad
      \snv \\
    \redva{\Delta}{\ind{\Sigma}(\tau_1)}{\alpha} \quad &= \\
      \{ \, v_1 \in \snv \; &| \;
        (\forall v_2. \; v_1 \reducestov \ret{v_2}
          \; \implies \; v_2 \in \redv{\cempty}{\tau_1}) \\
        &\wedge \; (\forall op
          \types \tau_2 \rightarrowtail \tau_3 \in \Sigma, v_3, v_4. \quad
          v_1 \reducestov \indintro{op}{v_3}{v_4} \\
          &\quad \implies v_3 \in \redv{\cempty}{\tau_2} \\
          &\qquad \quad \wedge \; (\forall v_5 \in \snv. \;
                   (\forall v_6. \; v_5 \reducestov \gbox{v_6} \; \implies \; v_6 \in \redv{\cempty}{\tau_3}) \\
                   &\qquad \qquad \implies \; \exists \beta < \alpha.
                     \quad \app{v_4}{v_5} \in \redca{\Delta}{\ind{\Sigma}(\tau_1)}{\beta})) \, \} \\
    \redca{\Delta}{\tau}{\alpha} \quad &= \\
      \{ \, c_1 \in \snc \; &| \;
        (\forall v_1. \; c_1 \reducestoc \return{v_1}
          \; \implies \; v_1 \in \redva{\Delta}{\tau}{\alpha}) \\
        &\wedge \; (\forall \loc, op, v_2. \quad
            c_1 \reducestoc \perform{\loc}{op}{v_2} \\
            &\quad \implies \;
            \exists \Sigma, \tau_2, \tau_3. \; \loc \types \Sigma \in \Delta
                            \; \wedge \; op \types \tau_2 \rightarrowtail \tau_3 \in \Sigma
                            \; \wedge \; v_2 \in \redv{\cempty}{\tau_2}) \\
        &\wedge \; (\forall \loc, op, v_3, c_2. \quad
           c_1 \reducestoc \letv{x}{\perform{\loc}{op}{v_3}}{c_2} \\
            &\quad \implies \;
            \exists \Sigma, \tau_4, \tau_5. \; \loc \types \Sigma \in \Delta
                            \; \wedge \; op \types \tau_4 \rightarrowtail \tau_5 \in \Sigma
                            \; \wedge \; v_3 \in \redv{\cempty}{\tau_4} \\
                            &\qquad \quad \wedge \; (\forall v_4 \in \snv. \;
                              (\forall v_5. \; v_4 \reducestov \gbox{v_5} \; \implies \; v_5 \in \redv{\cempty}{\tau_4}) \\
                              &\qquad \qquad \implies \; \exists \beta < \alpha.
                                 \quad \sbst{c_2}{x}{v_4} \in \redca{\Delta}{\tau}{\beta})) \, \}
\end{align*}
\caption{Reducibility predicates}
\label{fig:reducibility-predicates}
\end{figure}

\subsection{Properties of reducibility}

We prove the three properties of reducibility refered to as CR1, CR2 and
CR3 in Girard~\cite{girard1989proofs}.

Reducible values and computations are strongly normalizing (CR1):
\begin{lemma}
\label{lemma:red-implies-sn}
\begin{align*}
  &\redv{\Delta}{\tau} \subseteq \snv \\
  &\redc{\Delta}{\tau} \subseteq \snc
\end{align*}
\end{lemma}
\begin{proof}
We prove this by induction on $\tau$. Most of the cases follow directly
by the definitions of $\redvname$ and $\redcname$. The exceptions are
pair and function types.
\begin{description}
\item[$\tau_1 \times \tau_2$:] We have $\fst{v} \in \snv$ and
  $\snd{v} \in \snv$ and must show that $v \in \snv$. This holds because
  if there was an infinite chain of reductions for $v$ then there would
  be an infinite chain of reductions for $\fst{v}$ by congurence.
\item[$\tau_1 \arr \tau_2$:] We have
  $\forall v_2 \in \redv{\Delta}{\tau_1}. \; \app{v_1}{v_2} \in \snc$
  and must show that $v_1 \in \snv$. This holds because if there was an
  infinite chain of reductions for $v_1$ then there would be an infinite
  chain of reductions for $\app{v_1}{v_2}$ by congruence.
\end{description}
\end{proof}

Reduciblity predicates are compatible with reductions (CR2):
\begin{lemma}
\label{lemma:red-reduces}
\begin{align*}
  &v_1 \in \redva{\Delta}{\tau}{\alpha} \; \wedge \; v_1 \reducestov v_2 \;
    \implies \; v_2 \in \redva{\Delta}{\tau}{\alpha} \\
  &c_1 \in \redva{\Delta}{\tau}{\alpha} \; \wedge \; c_1 \reducestoc c_2 \;
    \implies \; c_2 \in \redva{\Delta}{\tau}{\alpha}
\end{align*}
\end{lemma}
\begin{proof}
  We prove this by induction on $\tau$. The reasoning for computations
  is the same for all types so we treat it as a separate case.
\begin{description}
\item[$1$:] If $v_1 \in \snv$ then $v_2 \in \snv$ because an infinite
  chain in $v_2$ would produce an infinite chain in $v_1$.
\item[$\tau_1 \times \tau_2$:] By congruence we have
  $\fst{v_1} \reducestov \fst{v_2}$ and
  $\snd{v_1} \reducestov \snd{v_2}$, so
  $\fst{v_2} \in \redv{\Delta}{\tau_1}$ and
  $\snd{v_2} \in \redv{\Delta}{\tau_2}$ follow by the induction
  hypothesis.
\item[$\tau_1 + \tau_2$:] Given $v_2 \reducestov \inl{v_3}$ we have
  $v_1 \reducestov \inl{v_3}$ and thus $v_3 \in
  \redv{\Delta}{\tau_1}$. Similarly, given $v_2 \reducestov \inr{v_4}$
  we have $v_4 \in \redv{\Delta}{\tau_2}$.
\item[$\tau_1 \arr \tau_2$:] Given $v_3 \in \redv{\Delta}{\tau_1}$, we
  have that $\app{v_1}{v_3}$ reduces to $\app{v_2}{v_3}$ by congruence,
  so $\app{v_2}{v_3} \in \redc{\Delta}{\tau_2}$ by the induction
  hypothesis.
\item[$\glob \tau$:] Given $v_2 \reducestov \gbox{v_3}$ we have
  $v_1 \reducestov \gbox{v_3}$ and thus $v_3 \in \redv{\cempty}{\tau}$.
\item[$\yoneda{\tau}$:] If $v_1 \in \snv$ then $v_2 \in \snv$ because an infinite
  chain in $v_2$ would produce an infinite chain in $v_1$.
\item[$\ind{\Sigma}(\tau)$:] Given $v_2 \reducestov \ret{v_3}$
  or $v_2 \reducestov \indintro{op}{v_4}{v_5}$ then $v_1$ also reduces to those
  values and thus they have the required properties.
\item[$\redcname$:] Given $c_2 \reducestoc \return{v_3}$,
  $c_2 \reducestoc \perform{\loc}{op}{v_4}$ or
  $c_2 \reducestoc \letv{x}{\perform{\loc}{op}{v_5}}{c_3}$ then $c_1$
  also reduces to those computations and thus they have the required
  properties.
\end{description}
\end{proof}

A term is neutral if it is not an introduction form. We consider
$\return{v}$, $\perform{\loc}{op}{v}$ and
$\letv{x}{\perform{\loc}{op}{v}}{c}$ to be the introduction forms for
computations. If a term is neutral, and whenever we reduce t we obtain a
term that is reducible, then the original term is reduceable (CR3).
\begin{lemma}
\label{lemma:red-neutral}
For all neutral values $v_n$ and neutral computations $c_n$:
\begin{align*}
  &(\forall v_r. \; v_n \stepstov v_r \implies v_r \in \redva{\Delta}{\tau}{\alpha}) \implies v_n \in \redva{\Delta}{\tau}{\alpha} \\
  &(\forall c_r. \; c_n \stepstoc c_r \implies c_r \in \redca{\Delta}{\tau}{\alpha}) \implies c_n \in \redva{\Delta}{\tau}{\alpha}
\end{align*}
\end{lemma}
\begin{proof}
  We prove this by induction on $\tau$. The reasoning for computations
  is the same for all types so we treat it as a separate case.
\begin{description}
\item[$1$:] Any chain from $v_n$ must pass through one of the $v_r$ and
  so must be finite.
\item[$\tau_1 \times \tau_2$:] As $v_n$ is neutral, the only reduction
  rule that applies to $\fst{v_n}$ is congruence, so any reduction must
  be of the form $\fst{v_n} \stepstov \fst{v_r}$. We have
  $v_r \in \redva{\Delta}{\tau_1 \times \tau_2}{\alpha}$ and thus
  $\fst{v_r} \in \redv{\Delta}{\tau_1}$, so by the induction hypothesis
  we have $\fst{v_n} \in \redv{\Delta}{\tau_1}$. Similarly, we have
  $\snd{v_n} \in \redv{\Delta}{\tau_2}$.
\item[$\tau_1 + \tau_2$:] Any reduction $v_n \reducestov \inl{v_i}$ must
  go through one of the $v_r$ because $v_n$ cannot equal $\inl{v_i}$ as
  it is neutral. From $v_r \reducestov \inl{v_i}$ we have
  $v_i \in \redv{\Delta}{\tau_1}$. Similarly we have that if
  $v_n \reducestov \inl{v_i}$ then $v_i \in \redv{\Delta}{\tau_2}$.
\item[$\tau_1 \arr \tau_2$:] For any
  $v_a \in \redv{\Delta}{\tau_1}$, we have by
  CR1(\ref{lemma:red-implies-sn}) that $v_a \in \snv$.
  We prove that $\app{v_n}{v_a} \in \redv{\Delta}{\tau_2}$ by
  induction on $\bound{v_a}$.
  By the outer induction hypothesis it is sufficient to show that any reduct
  of $\app{v_n}{v_a}$ is itself in $\redv{\Delta}{\tau_2}$.
  As $v_n$ is neutral the only reduction rules that apply to $\app{v_n}{v_a}$ are the
  congruences, so any reduction must be one of:
  \begin{itemize}
    \item $\app{v_n}{v_a} \stepstov \app{v_r}{v_a}$
    \item $\app{v_n}{v_a} \stepstov \app{v_n}{v_b}$ where $v_a \stepstov v_b$
  \end{itemize}
  In the first case, we have
  $v_r \in \redva{\Delta}{\tau_1 \arr \tau_2}{\alpha}$ and thus
  $\app{v_r}{v_a} \in \redv{\Delta}{\tau_2}$. In the second case, we
  have $v_b \in \redv{\Delta}{\tau_1}$ by CR2(\ref{lemma:red-reduces}),
  and $\bound{v_b} < \bound{v_a}$, so we have
  $\app{v_n}{v_b} \in \redv{\Delta}{\tau_2}$ by the inner induction
  hypothesis.
\item[$\glob \tau$:] Any reduction $v_n \reducestov \gbox{v_b}$ must
  go through one of the $v_r$ because $v_n$ cannot equal $\gbox{v_b}$ as
  it is neutral. From $v_r \reducestov \gbox{v_b}$ we have
  $v_b \in \redv{\Delta}{\tau}$.
\item[$\yoneda{\tau}$:] Any chain from $v_n$ must pass through one of
  the $v_r$ and so must be finite.
\item[$\ind{\Sigma}(\tau)$:] Any reduction $v_n \reducestov \ret{v_t}$
  or $v_n \reducestov \indintro{op}{v_p}{v_a}$ must go through one of
  the $v_r$ because $v_n$ cannot equal $\ret{v_t}$ or
  $\indintro{op}{v_p}{v_a}$ as it is neutral. Since they reduce from
  $v_r$ and $v_r \in \redva{\Delta}{\ind{\Sigma}(\tau)}{\alpha}$ they
  must have the required properties.
\item[$\redcname$:] Any reduction $c_n \reducestoc \return{v_t}$,
  $c_n \reducestoc \perform{\loc}{op}{v_p}$ or
  $c_n \reducestoc \letv{x}{\perform{\loc}{op}{v_p}}{c_k}$ must go
  through one of the $c_r$ because $c_n$ cannot equal $\return{v_t}$,
  $\perform{\loc}{op}{v_p}$ or $\letv{x}{\perform{\loc}{op}{v_p}}{c_k}$
  as it is neutral. Since they reduce from $c_r$ and
  $c_r \in \redca{\Delta}{\tau}{\alpha}$ they must have the required
  properties.
\end{description}
\end{proof}

As a corollary, we get that any neutral normal form is reducible. In
particular, the variable term $x$ is reducible.

\subsection{Effect weakening}

Reducibility can have its effect context weakened:
\begin{lemma}
\label{lemma:red-weaken}
\begin{align*}
  \redva{\Delta}{\tau}{\alpha} \subseteq \redva{\lbind{\Delta}{\loc}{\Sigma}}{\tau}{\alpha}
\end{align*}
\end{lemma}
\begin{proof}
  We prove this by induction on $\tau$ and $\alpha$. All the value cases
  follow immediately by the induction hypotheses as they do not mention
  $\Delta$ otherwise. The computation case follows by the induction
  hypothesis and the fact that context membership is preserved by
  weakening.
\end{proof}

\subsection{Compositionality of reducibility}

Next we show that the various term formers produce reducible terms from reducible subterms.

\subsubsection{Products}

Projection terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-proj}
\begin{align*}
  &v \in \redv{\Delta}{\tau_1 \times \tau_2} \; \implies \; \fst{v} \in \redv{\Delta}{\tau_1} \\
  &v \in \redv{\Delta}{\tau_1 \times \tau_2} \; \implies \; \snd{v} \in \redv{\Delta}{\tau_2}
\end{align*}
\end{lemma}
\begin{proof}
  Follows directly from the definition of $\redv{\Delta}{\tau_1 \times \tau_2}$
\end{proof}

Pair terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-pair}
\begin{align*}
  v_1 \in \redv{\Delta}{\tau_1} \; \wedge \; v_2 \in \redv{\Delta}{\tau_2}
  \; \implies \; (v_1, v_2) \in \redv{\Delta}{\tau_1 \times \tau_2}
\end{align*}
\end{lemma}
\begin{proof}
  By induction on $\bound{v_1} + \bound{v_2}$.  $\fst{(v_1, v_2)}$ is
  neutral so we can use CR3(\ref{lemma:red-neutral}). There are three
  possibilities for a reduct of $\fst{(v_1, v_2)}$.
  \begin{itemize}
  \item It can reduce by the $\fstname$ reduction rule to $v_1$.
  \item It can reduce by congruence to $\fst{(v_3, v_2)}$ where $v_1 \stepstov v_3$.
  \item It can reduce by congruence to $\fst{(v_1, v_4)}$ where $v_2 \stepstov v_4$.
  \end{itemize}
  The first case is reducible by hypothesis. The second and third cases
  hold by the inductive hypothesis using
  CR2(\ref{lemma:red-reduces}). The same reasoning applies to
  $\snd{(v_1, v_2}$.
\end{proof}

\subsubsection{Functions}

Application terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-app}
\begin{align*}
  v_1 \in \redv{\Delta}{\tau_1 \arr \tau_2} \; \wedge \; v_2 \in \redv{\Delta}{\tau_1}
  \; \implies \; \app{v_1}{v_2} \in \redc{\Delta}{\tau_2}
\end{align*}
\end{lemma}
\begin{proof}
  Follows directly from the definition of $\redv{\Delta}{\tau_1 \arr \tau_2}$
\end{proof}

Abstraction terms are reducible when built from a reducible subterm.
\begin{lemma}
\label{lemma:red-abs}
\begin{align*}
  (\forall v_a \in \redv{\Delta}{\tau_1}. \; \sbst{c_b}{x}{v_a} \in \redc{\Delta}{\tau_2})
  \; \implies \; \abs{x}{c_b} \in \redv{\Delta}{\tau_1 \arr \tau_2}
\end{align*}
\end{lemma}
\begin{proof}
  We need to show that
  $\app{\abs{x}{c_b}}{v_a} \in \redc{\Delta}{\tau_2}$ for any
  $v_a \in \redv{\Delta}{\tau_1}$. We proceed by induction on
  $\bound{c_b} + \bound{v_a}$. As $\app{\abs{x}{c_b}}{v_a}$ is neutral,
  we can use CR3(\ref{lemma:red-neutral}). There are three possibilities for a
  reduct of $\app{\abs{x}{c_b}}{v_a}$.
  \begin{itemize}
  \item It can reduce by beta reduction to $\sbst{c_b}{x}{v_a}$.
  \item It can reduce by congruence to $\app{\abs{x}{c_d}}{v_a}$ where $c_b \stepstoc c_d$.
  \item It can reduce by congruence to $\app{\abs{x}{c_b}}{v_e}$ where $v_a \stepstov v_e$.
  \end{itemize}
  The first case is reducible by hypothesis. The second and third cases
  are reducible by the inductive hypothesis using
  CR2(\ref{lemma:red-reduces}).
\end{proof}

\subsubsection{Sums}

Injection terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-inj}
\begin{align*}
  &v_i \in \redv{\Delta}{\tau_1} \; \implies \; \inl{v_i} \in \redv{\Delta}{\tau_1 + \tau_2} \\
  &v_i \in \redv{\Delta}{\tau_2} \; \implies \; \inr{v_i} \in \redv{\Delta}{\tau_1 + \tau_2}
\end{align*}
\end{lemma}
\begin{proof}
  By induction on $\bound{v_i}$. For left injection, we must show two
  things:
  \begin{itemize}
  \item We must show that $\inl{v_i} \in \snv$. The only reduction rule that can apply
    to $\inl{v_i}$ is the congruence rule. Thus any reduct would be of
    the form $\inl{v_j}$ with $v_i \stepstov v_j$ and so, by the
    inductive hypothesis via CR2(\ref{lemma:red-reduces}), would be
    strongly normalizing. Since every reduct is strongly normalizing then
    so is $\inl{v_i}$.
  \item We must show that $v_j \in \redv{\Delta}{\tau_1}$ for any $v_j$ such that $\inl{v_i} \reducestov \inl{v_j}$.
    The reduction is either from reflexivity or by stepping to some intermediate term. If
    it is by reflexivity then $v_j$ is reducible by hypothesis. Only the
    congruence reduction rule applies to $\inl{v_i}$, so the intermediate
    term must be of the form $\inl{v_k}$, which means we can use the
    inductive hypothesis on $v_k$ CR2(\ref{lemma:red-reduces}) to show
    that $v_j$ is reducible.
  \end{itemize}

  The case for right injection follows by the same reasoning.
\end{proof}

Case terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-case}
\begin{align*}
  &v_1 \in \redv{\Delta_1}{\tau_1 + \tau_2} \\
  &\wedge \; (\forall v_4 \in \redv{\Delta_1}{\tau_1}. \; \sbst{v_2}{x_1}{v_4} \in \redv{\Delta_2}{\tau_3}) \\
  &\wedge \; (\forall v_5 \in \redv{\Delta_1}{\tau_2}. \; \sbst{v_3}{x_2}{v_5} \in \redv{\Delta_2}{\tau_3}) \\
  &\implies \; \case{v_1}{x_1}{v_2}{x_2}{v_3} \in \redv{\Delta_2}{\tau_3}
\end{align*}
\end{lemma}
\begin{proof}
  By induction on $\bound{v_1} + \bound{v_2} + \bound{v_3}$.
  $\case{v_1}{x_1}{v_2}{x_2}{v_3}$ is neutral so we can use CR3(\ref{lemma:red-neutral}). There are five
  possibilities for a reduct of $\case{v_1}{x_1}{v_2}{x_2}{v_3}$.
  \begin{itemize}
  \item It can reduce to $\sbst{v_2}{x_1}{v_4}$ if $v_1$ is of the form $\inl{v_4}$.
  \item It can reduce to $\sbst{v_3}{x_2}{v_5}$ if $v_1$ is of the form $\inr{v_5}$.
  \item It can reduce by congruence to $\case{v_6}{x_1}{v_2}{x_2}{v_3}$ where $v_1 \stepstov v_6$.
  \item It can reduce by congruence to $\case{v_1}{x_1}{v_7}{x_2}{v_3}$ where $v_2 \stepstov v_7$.
  \item It can reduce by congruence to $\case{v_1}{x_1}{v_2}{x_2}{v_8}$ where $v_3 \stepstov v_8$.
  \end{itemize}
  In the first case we have $v_4 \in \redv{\Delta_1}{\tau_1}$ from $v_1 \in \redv{\Delta_1}{\tau_1 + \tau_2}$ 
  and so the reduct is reducible by hypothesis. Similarly, in the second case we have $v_5 \in \redv{\Delta_1}{\tau_2}$
  and so the reduct is reducible by hypothesis. The last three cases
  are reducible by the inductive hypothesis using
  CR2(\ref{lemma:red-reduces}).
\end{proof}

\subsubsection{Unit}

Unit terms are reducible.
\begin{lemma}
\label{lemma:red-unit}
\begin{align*}
  () \in \redv{\Delta}{1}
\end{align*}
\end{lemma}
\begin{proof}
  No reduction rules apply to $()$ so it is strongly normalizable.
\end{proof}

\subsubsection{Global modality}

Boxing terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-box}
\begin{align*}
  &v_b \in \redv{\cempty}{\tau} \; \implies \; \gbox{v_b} \in \redv{\Delta}{\tau}
\end{align*}
\end{lemma}
\begin{proof}
  By induction on $\bound{v_b}$. We must show two things:
  \begin{itemize}
  \item We must show that $\gbox{v_b} \in \snv$. The only reduction rule that can apply
    to $\gbox{v_b}$ is the congruence rule. Thus any reduct would be of
    the form $\gbox{v_c}$ with $v_b \stepstov v_c$ and so, by the
    inductive hypothesis via CR2(\ref{lemma:red-reduces}), would be
    strongly normalizing. Since every reduct is strongly normalizing then
    so is $\gbox{v_b}$.
  \item We must show that $v_c \in \redv{\cempty}{\tau}$ for any $v_c$
    such that $\gbox{v_b} \reducestov \gbox{v_c}$. The reduction is
    either from reflexivity or by stepping to some intermediate term. If
    it is by reflexivity then $v_c$ is reducible by hypothesis. Only the
    congruence reduction rule applies to $\gbox{v_b}$, so the
    intermediate term must be of the form $\gbox{v_d}$, which means we
    can use the inductive hypothesis on $v_d$ via
    CR2(\ref{lemma:red-reduces}) to show that $v_c$ is reducible.
  \end{itemize}
\end{proof}

Unboxing terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-unbox}
\begin{align*}
  &v_1 \in \redv{\Delta_1}{\glob \tau_1} \\
  &\wedge \; (\forall v_3 \in \redv{\cempty}{\tau_1}.
    \; \sbst{v_2}{x}{v_3} \in \redv{\Delta_2}{\tau_2}) \\
  &\implies \; \gunbox{x}{v_1}{v_2} \in \redv{\Delta_2}{\tau_2}
\end{align*}
\end{lemma}
\begin{proof}
  By induction on $\bound{v_1} +
  \bound{v_2}$. $\gunbox{x}{v_1}{v_2}$ is neutral so we can use
  CR3(\ref{lemma:red-neutral}). There are three possibilities for a
  reduct of $\gunbox{x}{v_1}{v_2}$:
  \begin{itemize}
  \item It can reduce to $\sbst{v_2}{x}{v_3}$ if $v_1$ is of the form $\gbox{v_3}$.
  \item It can reduce by congruence to $\gunbox{x}{v_4}{v_2}$ where $v_1 \stepstov v_4$.
  \item It can reduce by congruence to $\gunbox{x}{v_1}{v_5}$ where $v_2 \stepstov v_5$.
  \end{itemize}
  In the first case we have $v_3 \in \redv{\cempty}{\tau_1}$ from
  $v_1 \in \redv{\Delta_1}{\glob \tau_1}$ and so the reduct is reducible
  by hypothesis. The second and third cases are reducible by the
  inductive hypothesis using CR2(\ref{lemma:red-reduces}).
\end{proof}

\subsubsection{$\yonedaname$ types}

Selection terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-out}
\begin{align*}
  &v_s \in \redv{\Delta}{\yoneda{\Sigma_1 + \Sigma_2}}
    \; \implies \; \outl{v_s} \in \redv{\Delta}{\yoneda{\Sigma_1}}
  &v_s \in \redv{\Delta}{\yoneda{\Sigma_1 + \Sigma_2}}
    \; \implies \; \outr{v_s} \in \redv{\Delta}{\yoneda{\Sigma_2}}
\end{align*}
\end{lemma}
\begin{proof}
  By induction on $\bound{v_s}$. The only reduction rule that can apply
  to $\outl{v_s}$ is the congruence rule. Thus any reduct would be of
  the form $\outl{v_r}$ with $v_s \stepstov v_r$ and so, by the
  inductive hypothesis via CR2(\ref{lemma:red-reduces}), would be
  strongly normalizing. Since every reduct is strongly normalizing then
  so is $\outl{v_s}$.

  The case for right selection follows by the same reasoning.
\end{proof}

\subsubsection{Computations}

Return terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-return}
\begin{align*}
  &v_r \in \redv{\Delta}{\tau} \\
  &\implies \; \return{v_r} \in \redc{\Delta}{\tau}
\end{align*}
\end{lemma}
\begin{proof}
  By induction on $\bound{v_r}$. We must show three things:
  \begin{itemize}
  \item We must show that $\return{v_r} \in \snc$. The only reduction rule that can apply
    to $\return{v_r}$ is the congruence rule. Thus any reduct would be of
    the form $\return{v_t}$ with $v_r \stepstov v_t$ and so, by the
    inductive hypothesis via CR2(\ref{lemma:red-reduces}), would be
    strongly normalizing. Since every reduct is strongly normalizing then
    so is $\return{v_r}$.
  \item We must show that $v_t \in \redv{\Delta}{\tau}$ for any $v_t$
    such that $\return{v_r} \reducestov \return{v_t}$. The reduction is
    either from reflexivity or by stepping to some intermediate term. If
    it is by reflexivity then $v_t$ is reducible by hypothesis. Only the
    congruence reduction rule applies to $\return{v_r}$, so the
    intermediate term must be of the form $\return{v_s}$, which means we
    can use the inductive hypothesis on $v_s$ via
    CR2(\ref{lemma:red-reduces}) to show that $v_t$ is reducible.
  \item We must show that various propertes hold for any term of the
    form $\perform{\loc}{\Sigma}{v_a}$ or
    $\letv{x}{\perform{\loc}{op}{v_a}}{c_k}$ that $\return{v_r}$ reduces
    to. The reduction would have to be by stepping to some intermediate
    term as it can't be via reflexivity. Only the congruence reduction
    rule applies to $\return{v_r}$, so the intermediate term must be of
    the form $\return{v_t}$, which means we can use the inductive
    hypothesis on $v_t$ via CR2(\ref{lemma:red-reduces}) to show that
    the properties would hold.
  \end{itemize}
\end{proof}

Non-neutral perform terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-perform-intro}
\begin{align*}
  &\loc_e \types \Sigma \in \Delta
   \; \wedge \; op \types \tau_p \rightarrowtail \tau_a \in \Sigma
   \; \wedge \; v_p \in \redv{\cempty}{\tau_p} \\
  &\implies \; \perform{\loc_e}{op}{v_p} \in \redc{\Delta}{\glob \tau_a}
\end{align*}
\end{lemma}
\begin{proof}
  By induction on $\bound{v_p}$. We must show three things:
  \begin{itemize}
  \item We must show that $\perform{\loc_e}{op}{v_p} \in \snc$. The only
    reduction rule that can apply to $\perform{\loc_e}{op}{v_p}$ is the
    congruence rule. Thus any reduct would be of the form
    $\perform{\loc_e}{op}{v_q}$ with $v_p \stepstov v_q$ and so, by the
    inductive hypothesis via CR2(\ref{lemma:red-reduces}), would be
    strongly normalizing. Since every reduct is strongly normalizing
    then so is $\perform{\loc_e}{op}{v_p}$.
  \item We must show that $v_q \in \redv{\cempty}{\tau_p}$ for any $v_q$
    such that
    $\perform{\loc_e}{op}{v_p} \reducestov \perform{\loc_e}{op}{v_q}$. The
    reduction is either from reflexivity or by stepping to some
    intermediate term. If it is by reflexivity then $v_q$ is reducible
    by hypothesis. Only the congruence reduction rule applies to
    $\perform{\loc_e}{op}{v_p}$, so the intermediate term must be of the form
    $\perform{\loc_e}{op}{v_r}$, which means we can use the inductive hypothesis on
    $v_r$ via CR2(\ref{lemma:red-reduces}) to show that $v_q$ is
    reducible.
  \item We must show that various propertes hold for any term of the
    form $\return{v_r}$, $\perform{\loc_f}{\Sigma}{v_q}$ (where
    $\loc_f \neq \loc_e$) or $\letv{x}{\perform{\loc}{op}{v_p}}{c_k}$
    that $\perform{\loc_e}{op}{v_p}$ reduces to. The reduction would
    have to be by stepping to some intermediate term as it can't be via
    reflexivity. Only the congruence reduction rule applies to
    $\perform{\loc_e}{op}{v_p}$, so the intermediate term must be of the form
    $\perform{\loc_e}{op}{v_q}$, which means we can use the inductive hypothesis on
    $v_q$ via CR2(\ref{lemma:red-reduces}) to show that the properties
    would hold.
  \end{itemize}
\end{proof}

Non-neutral bind terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-bind-intro}
\begin{align*}
  &\loc_e \types \Sigma \in \Delta
   \; \wedge \; op \types \tau_p \rightarrowtail \tau_a \in \Sigma
   \; \wedge \; v_p \in \redv{\Delta}{\tau_p} \\
  &\wedge \; (\forall v_a \in \redva{\Delta}{\glob \tau_a}. \; \sbst{c_k}{x}{v_a} \in \redc{\Delta}{\tau_k}) \\
  &\implies \; \letv{x}{\perform{\loc_e}{op}{v_p}}{c_k} \in \redc{\Delta}{\tau_k}
\end{align*}
\end{lemma}
\begin{proof}
  By induction on $\bound{v_p} + \bound{c_k}$. We must show three things:
  \begin{itemize}
  \item We must show that $\letv{x}{\perform{\loc_e}{op}{v_p}}{c_k} \in \snc$. The only
    reduction rule that can apply to $\letv{x}{\perform{\loc_e}{op}{v_p}}{c_k}$ is the
    congruence rule. Thus any reducts would be one of two forms:
    \begin{itemize}
    \item $\letv{x}{\perform{\loc_e}{op}{v_q}}{c_k}$ with $v_q \stepstov v_q$
    \item $\letv{x}{\perform{\loc_e}{op}{v_p}}{c_j}$ with $c_k \stepstoc c_j$
    \end{itemize}
    in both cases the reduct is strongly normalizing by the inductive
    hypothesis via CR2(\ref{lemma:red-reduces}). Since every reduct is
    strongly normalizing then so is
    $\letv{x}{\perform{\loc_e}{op}{v_p}}{c_k}$.
  \item We must show that $v_q \in \redv{\cempty}{\tau_p}$ and
    $forall v_a \in \redva{\Delta}{\glob \tau_a}. \; \sbst{c_j}{x}{v_a}
    \in \redc{\Delta}{\tau_k}$ for any $v_q$ and $c_j$ such that
    $\letv{x}{\perform{\loc_e}{op}{v_p}}{c_k} \reducestov
    \letv{x}{\perform{\loc_e}{op}{v_q}}{c_j}$. The reduction is either
    from reflexivity or by stepping to some intermediate term. If it is
    by reflexivity then $v_q$ and $c_j$ are reducible by
    hypothesis. Only the congruence reduction rule applies to
    $\letv{x}{\perform{\loc_e}{op}{v_p}}{c_k}$, so the intermediate term
    must be of the form $\letv{x}{\perform{\loc_e}{op}{v_r}}{c_l}$,
    which means we can use the inductive hypothesis on $v_r$ and $c_l$
    via CR2(\ref{lemma:red-reduces}) to show that $v_q$ and $c_j$ are
    reducible.
  \item We must show that various propertes hold for any term of the
    form $\return{v_r}$, $\perform{\loc_g}{\Sigma}{v_p}$ or
    $\letv{x}{\perform{\loc_f}{op}{v_q}}{c_j}$ (where
    $\loc_f \neq \loc_e$) that
    $\letv{x}{\perform{\loc_f}{op}{v_p}}{c_k}$ reduces to. The reduction
    would have to be by stepping to some intermediate term as it can't
    be via reflexivity. Only the congruence reduction rule applies to
    $\letv{x}{\perform{\loc_f}{op}{v_p}}{c_k}$, so the intermediate term
    must be of the form $\letv{x}{\perform{\loc_f}{op}{v_r}}{c_l}$,
    which means we can use the inductive hypothesis on $v_r$ and $c_l$
    via CR2(\ref{lemma:red-reduces}) to show that the properties would
    hold.
  \end{itemize}
\end{proof}

Perform terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-perform}
\begin{align*}
  &\loc_e \types \Sigma \in \Delta
   \; \wedge \; op \types \tau_p \rightarrowtail \tau_a \in \Sigma \\
  &v_l \in \redv{\Delta}{\yoneda{\Sigma}}
  \; \wedge \; v_a \in \redv{\Delta}{\tau_p} \\
  &\implies \; \perform{v_l}{op}{v_a} \in \redc{\Delta}{\glob \tau_a}
\end{align*}
\end{lemma}
\begin{proof}
\todo{Write out proof}
\end{proof}

Bind terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-bind}
\begin{align*}
  &c_1 \in \redca{\Delta}{\tau_1}{\alpha} \\
  &\wedge \; (\forall v \in \redva{\cempty}{\tau_1}{\alpha}. \; \sbst{c_2}{x}{v} \in \redc{\Delta}{\tau_2}) \\
  &\implies \; \letv{x}{c_1}{c_2} \in \redc{\Delta}{\tau_2}
\end{align*}
\end{lemma}
\begin{proof}
  By induction on $\alpha$ and $\bound{c_1} + \bound{c_2}$.
  \todo{Write out proof}
\end{proof}

Run terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-run}
\begin{align*}
  &c_r \in \redc{\cempty}{\tau} \\
  &\implies \; \grun{c_r} \in \redv{\Delta}{\tau}
\end{align*}
\end{lemma}
\begin{proof}
\todo{Write out proof}
\end{proof}

Reification terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-reify}
\begin{align*}
  &c \in \redca{\lbind{\Delta}{\loc}{\Sigma}}{\glob \tau}{\alpha} \\
  &\implies \; \reify{\loc}{c} \in \redca{\Delta}{\ind{\Sigma}{\tau}}{\alpha}
\end{align*}
\end{lemma}
\begin{proof}
\todo{Write out proof}
\end{proof}

\subsubsection{Inductive types}

Recursion terms are reducible when built from reducible subterms.
\begin{lemma}
\label{lemma:red-rec}
\begin{align*}
  &v_s \in \redva{\Delta_1}{\ind{\{ op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1};
      \ldots; op_n\types \tau_{p_n} \rightarrowtail \tau_{a_n} \}}(\tau_1)}{\alpha} \\
  &\wedge \; (\forall v_r \in \redva{\cempty}{\tau_1}{\alpha}. \\
    &\qquad \sbst{v_e}{x_r}{v_r} \in \redva{\Delta_2}{\tau_2}{\alpha}) \\
  &\wedge \; (\forall v_{p_1} \in \redva{\cempty}{\tau_{p_1}}{\alpha}, \;
      v_{k_1} \in \redva{\Delta_1}{\glob \tau_{a_1} \arr \tau_2}{\alpha}. \\
    &\qquad \sbst{\sbst{v_1}{x_{p_1}}{v_{p_1}}}{x_{k_1}}{v_{k_1}} \in \redva{\Delta_2}{\tau_2}{\alpha}) \\
  &\ldots \\
  &\wedge \; (\forall v_{p_n} \in \redva{\cempty}{\tau_{p_n}}{\alpha}, \;
      v_{k_n} \in \redva{\Delta_1}{\glob \tau_{a_n} \arr \tau_2}{\alpha}. \\
    &\qquad \sbst{\sbst{v_n}{x_{p_n}}{v_{p_n}}}{x_{k_n}}{v_{k_n}} \in \redva{\Delta_2}{\tau_2}{\alpha}) \\
  &\implies \\
  &\indelimm{v_s}{\indretcase{x_r}{v_e} \seq
            \indcase{op_1}{x_{p_1}}{x_{k_1}}{v_1} \seq \ldots \seq
            \indcase{op_n}{x_{p_n}}{x_{k_n}}{v_n}}
    \in \redva{\Delta_2}{\tau_2}{\alpha}
\end{align*}
\end{lemma}
\begin{proof}
\todo{Write out proof}
\end{proof}

\subsection{Reducible substitutions}

We define a notion of reducibility for substitutions as a reducibility
predicate $\redsc{\Delta}{\Gamma}$ indexed by an effect context $\Delta$
and a typing context $\Gamma$. We write $\reds{\Gamma}$ for
$\redsc{\cempty}{\Gamma}$.

\todo{Write out definition}

\subsection{Reducibility theorem}
Well-typed terms are reducible under any reducible substitution
\begin{align*}
    &\Gamma \turnv v \types \tau
      \quad \wedge \quad \sigma \in \reds{\Gamma}
      \quad \implies \quad v[\sigma] \in \redv{\clocs{\Gamma}}{\tau} \\
    &\Gamma \turnc c \types \tau
      \quad \wedge \quad \sigma \in \reds{\Gamma}
      \quad \implies \quad c[\sigma] \in \redc{\clocs{\Gamma}}{\tau} \\
\end{align*}
\begin{proof}
\todo{Write out proof}
\end{proof}

As a corollary, since the substitution that replaces every variable with itself is reducible,
we get:
\begin{align*}
    &\Gamma \turnv v \types \tau
      \quad \implies \quad v \in \redv{\clocs{\Gamma}}{\tau} \\
    &\Gamma \turnc c \types \tau
      \quad \implies \quad c \in \redc{\clocs{\Gamma}}{\tau} \\
\end{align*}
from which immediately follows that well-typed terms are strongly normalizing.

\section{Embedding the Simply-Typed Lambda Calculus}
\label{apdx:embedding}
We define an embedding $\embed{\_}$ on types, contexts and terms of the
Simply-Typed Lambda Calculus (STLC) in Fig.\ref{fig:stlc}.
\begin{figure}
\begin{equation*}
\begin{aligned}[c]
  &\embed{1} = 1 \\
  &\embed{A \times B} = \embed{A} \times \embed{B} \\
  &\embed{A \rightarrow B} = \glob \embed{A} \arr \embed{B} \\ \\
  &\embed{\cempty} = \cempty \\
  &\embed{\lbind{\Gamma}{x}{A}} = \gbind{\embed{\Gamma}}{x}{\embed{A}}
\end{aligned}
\qquad
\begin{aligned}[c]
  &\embed{x} = x \\
  &\embed{(e, f)} = (\embed{e}, \embed{f}) \\
  &\embed{\fst{e}} = \fst{\embed{e}} \\
  &\embed{\snd{e}} = \snd{\embed{e}} \\
  &\embed{\abs{x}{e}} =
    \abs{(\gbox{x})}{\return{\embed{e}}} \\
  &\embed{\app{f}{e}} = \grun{\app{\embed{f}}{\embed{e}}}
\end{aligned}
\end{equation*}
\caption{Embedding of Simply-Typed Lambda Calculus into values}
\label{fig:stlc}
\end{figure}

Note that the embedding produces contexts with only global bindings so
that:
\begin{lemma}
\label{lemma:stlc-global-context}
  \begin{equation*}
    \cdiv{\embed{\Gamma}} = \embed{\Gamma}
  \end{equation*}
\end{lemma}

This embedding maps well-typed STLC terms to well-typed values:
\begin{lemma}
\label{lemma:stlc-well-typed}
  \begin{equation*}
    \Gamma \turnstlc e \types t
    \quad \implies \quad \embed{\Gamma} \turnv \embed{e} \types \embed{t}
  \end{equation*}
\end{lemma}
which is proved by induction on the STLC typing derivation. The case for
the app rule relies on \ref{lemma:stlc-global-context} in order to apply
$\grunname$ to the embedded expression.

Reduction of STLC terms is modelled by the reduction of the embedded
terms:
\begin{lemma}
\label{lemma:stlc-reduces}
  \begin{equation*}
    e_1 \stepstostlc e_2
    \quad \implies \quad \embed{e_1} \reducestov \embed{e_2}
  \end{equation*}
\end{lemma}
We prove this via induction on the STLC reduction relation. All cases
are trivial except for the function beta-reduction case:
\begin{equation*}
  \embed{\app{(\abs{x}{e_1})}{e_2}} \reducestov \embed{\sbst{e_1}{x}{e_2}}
\end{equation*}
which reduces to:
\begin{equation*}
\grun{\app{(\abs{(\gbox{x})}{\return{\embed{e_1}}})}{\embed{e_2}}}  \reducestov \embed{\sbst{e_1}{x}{e_2}}
\end{equation*}
applying the function beta rule, the $\grunname$ reduction rule and reflexivity this becomes:
\begin{equation*}
\embed{\sbst{e_1}{x}{e_2}} = \sbst{\embed{e_1}}{x}{\embed{e_2}}
\end{equation*}
which follows by induction on $e_1$.

Normal forms of STLC are embedded to normal forms:
\begin{lemma}
\label{lemma:stlc-reduces}
  \begin{equation*}
    e \stepsnotstlc
    \quad \implies \quad \embed{e} \stepsnotv
  \end{equation*}
\end{lemma}
This follows from observing that the neutral forms and normal forms of
STLC embed to neutral and normal forms respectively.

Since our langauge is strongly normalizing
(\ref{lemma:strong-normalization}) this is sufficient to show that the
embedding correctly models STLC.

\section{Denotational semantics}
\label{apdx:denotational}

In this section we prove that the denotatoinal semantics is sound and
adequate with respect to the operational semantics. We start by writing
out the full denotational semantcs in
Figs.~\ref{fig:denotational-types}, \ref{fig:denotational-values} and
\ref{fig:denotational-computations}.


where 

\begin{figure}
  \begin{subfigure}[T]{0.4\textwidth}
\begin{align*}
\sem{\tau} &\mathrel{:} \pshlawcop \\
\sem{1} &= 1 \\
\sem{\tau_1 \times \tau_2} &= \sem{\tau_1} \times \sem{\tau_2} \\
\sem{\tau_1 + \tau_2} &= \sem{\tau_1} + \sem{\tau_2} \\
\sem{\tau_1 \arr \tau_2} &= \arrow{\sem{\tau_1}}{\mon(\sem{\tau_2})} \\
\sem{\glob \tau} &= \glob \sem{\tau} \\
\sem{\yoneda{\Sigma}} &= \yoneda{\sem{\Sigma}} \\
\sem{\ind{\Sigma}(\tau)} = \ind{\sem{\Sigma}}(\sem{\tau})
\end{align*}
\end{subfigure}
\begin{subfigure}[T]{0.4\textwidth}
\begin{align*}
\sem{\Gamma} &\mathrel{:} \pshlawcop \\
\sem{\cempty} &= 1 \\
\sem{\lbind{\Gamma}{x}{\tau}} &= \sem{\Gamma} \times \sem{\tau} \\
\sem{\gbind{\Gamma}{x}{\tau}} &= \sem{\Gamma} \times \glob \sem{\tau} \\
\sem{\lbind{\Gamma}{\loc}{\Sigma}} &= \sem{\Gamma} \times \yoneda{\sem{\Sigma}} \\
\end{align*}
\begin{itemize}
\item $\sem{\Sigma}$ is the $\kappa$-bounded Lawvere theory corresponding to the
  signature $\Sigma$.
\end{itemize}
\end{subfigure}
\caption{Denotational semantics: Types and Contexts}
\label{fig:denotational-types}
\end{figure}

\begin{figure}
\begin{align*}
\sem{\Gamma \turnv v \types \tau} &\mathrel{:} \homset{\pshlawcop}{\sem{\Gamma}}{\sem{\tau}} \\
\sem{\lbind{\Gamma}{x}{\tau} \turnv x \types \tau} &= \snd \\
\sem{\gbind{\Gamma}{x}{\tau} \turnv x \types \tau} &= \counit{\glob}_{\sem{\tau}} \circ \snd \\
\sem{\lbind{\Gamma}{\loc}{\Sigma} \turnv \loc \types \yoneda{\Sigma}} &= \snd \\
\sem{\Gamma \turnv \gbox{v} \types \glob \tau} &=
  \glob \sem{\cdiv{\Gamma} \turnv v \types \tau}
  \circ \comult{\glob}_{\sem{\cdiv{\Gamma}}} \circ \bdiv{\Gamma} \\
\sem{\Gamma \turnv \gunbox{x}{v_1}{v_2} \types \tau_2} &=
  \sem{\gbind{\Gamma}{x}{\tau_1} \turnv v_2 \types \tau_2}
  \circ \langle id_{\sem{\Gamma}},
                \sem{\Gamma \turnv v_1 \types \glob \tau_1} \rangle \\
\sem{\Gamma \turnv \grun{c} \types \glob \tau} &=
  \ununit_{\sem{\tau}}
  \circ \glob \sem{\cdiv{\Gamma} \turnc c \types \tau}
  \circ \comult{\glob}_{\sem{\cdiv{\Gamma}}} \circ \bdiv{\Gamma}
\end{align*}
\begin{itemize}
\item $\counit{\glob}$ is the counit of $\glob$.
\item $\comult{\glob}$ is the comultiplication of $\glob$.
\item $\bdiv{\Gamma} \types \Gamma \rightarrow \cdiv{\Gamma}$ is the morphism
that projects the global bindings from $\Gamma$
\item $\ununit_P$ is the isomorphism from $\glob \mon(P)$ to $\glob P$.
\end{itemize}
\caption{Denotational semantics: Values}
\label{fig:denotational-values}
\end{figure}

\begin{figure}
\begin{align*}
\sem{\Gamma \turnc v \types \tau} &\mathrel{:} \homset{\pshlawcop}{\sem{\Gamma}}{\mon(\sem{\tau})} \\
\sem{\Gamma \turnc \return{v} \types \tau} &= \unit{\mon}_{\sem{\tau}} \circ \sem{\Gamma \turnv v \types \tau} \\
\sem{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2} &=
    \mult{\mon}_{\sem{\tau_2}}
    \circ \mon(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2})
    \circ \strength{\mon}_{\sem{\Gamma}, \sem{\tau_1}}
    \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnc c_1 \types \tau_1} \rangle \\
\sem{\Gamma \turnc \perform{v_1}{op}{v_2} \types \tau_a} &=
  op_{\mon}
       \circ \langle \sem{\Gamma \turnv v_1 \types \yoneda{\Sigma}},
                      \sem{\Gamma \turnv v_2 \types \tau_p} \rangle \\
\sem{\Gamma \turnc \reify{\loc}{c} \types \ind{\Sigma}(\glob \tau)} &=
  \reifyc{\Sigma}{\tau} \circ \currymap(\sem{\lbind{\Gamma}{\loc}{\Sigma} \turnc c \types \glob \tau}) \\
\end{align*}
\begin{itemize}
\item $\unit{\mon}$ is the unit of $\mon$.
\item $\mult{\mon}$ is the multiplication of $\mon$.
\item $\strength{\mon}$ is the strength of $\mon$.
\end{itemize}
\caption{Denotational semantics: Computations}
\label{fig:denotational-computations}
\end{figure}

\subsection{Soundness}
\subsection{Contextual equivalence}
\subsection{Compositionality}
\subsection{Adequacy}

\section{OxCaml implementation}
\label{apdx:implementation}
The implementation of our OxCaml effect reflection interface in terms of
unchecked effects\footnote{For convenience, we use the new handler
  syntax from OCaml 5.3. However, the most recently released version of
  OxCaml is based on OCaml 5.2 so the actual implementation uses
  \lstinline[style=oxcaml]{Effect.Deep.match_with} instead} is shown in
Fig.~\ref{fig:implementation}.
\begin{figure}
\begin{lstlisting}[style=oxcaml]
module Handler (E : Sig) = struct

  type t = { perform : 'r. 'r E.t -> 'r }

end

module Reflection (E : Sig) = struct
  open Term(E)
  open Handler(E)

  let reify f = exclave
    let module Eff =
      struct type 'a Effect.t += C : 'a E.t -> 'a Effect.t end
    in
    let handler = { perform = fun o -> Effect.perform (Eff.C o) } in
    match f handler with
    | v -> Return v
    | effect Eff.C o, k -> Perform(o, fun x -> Effect.continue k x)

  let reify_global f =
    let module Eff =
      struct type 'a Effect.t += C : 'a E.t -> 'a Effect.t end
    in
    let handler =
      { perform = fun o -> Effect.perform (Eff.C o) }
    in
    match f handler with
    | v -> Return v
    | effect Eff.C o, k -> Perform(o, fun x -> Effect.continue k x)

  let perform t =
   fun h -> h.perform t

end

module Select (L : Sig) (R : Sig) = struct
  open Sum(L)(R)

  let outl ({perform} : Handler(Sum(L)(R)).t) : Handler(L).t =
    exclave {perform = fun op -> perform (Left op)}

  let outr ({perform} : Handler(Sum(L)(R)).t) : Handler(R).t =
    exclave {perform = fun op -> perform (Right op)}

end
\end{lstlisting}
\caption{Implementation}
\label{fig:implementation}
\end{figure}

\subsection{More efficient implementation in terms of fibers}

In the implementation in Fig.\ref{fig:implementation}, because we are
using OCaml's existing algebraic effect support, whenever an operation
is reflected we perform a linear search up the stack of handlers until
we find the corresponding one. This search is not actually necessary:
there is a one-to-one relationship between the handlers and the handler
values used to reflect operations. The linear search could be eliminated
by extending the OxCaml runtime to support direct references to
handlers. By allowing local values to point directly to the fiber
structures that implement effect handlers, operations can jump
immediately to the correct handler without traversing the stack. This
optimization maintains the same interface while providing constant-time
operation dispatch. It is roughly the implementation technique described
by Ma et al.~\cite{ma2024lexical}.

\section{Effect reflection in practice}
\label{apdx:examples}
\subsection{Example: State}

As shown in the introduction we can implement an effect handler for
state using out effect reflection API (Fig.~\ref{fig:ex-state}).

\begin{figure}
\begin{lstlisting}[style=oxcaml]
  module State = struct
    type 'a t =
      | Get : int t
      | Set : int -> unit t
  end

  module State_eff = Reflection(State)
  type handler = Handler(State).t
  open Term(State)

  let handle_state i f =
    let rec loop (s : int) = function
      | Return x -> x
      | Perform(Get, k) -> loop s (State_eff.continue k s)
      | Perform(Set s', k) -> loop s' (State_eff.continue k ())
    in
    loop i (State_eff.reify f)

  let get h =
    State_eff.perform Get h

  let set h i =
    State_eff.perform (Set i) h
\end{lstlisting}
\caption{Example: State}
\label{fig:ex-state}
\end{figure}

The implementation has the following interface:
\begin{lstlisting}[style=oxcaml]
    type handler

    val handle_state : int -> (handler @ local -> 'a) @ local -> 'a
    val get : handler @ local -> int
    val set : handler @ local -> int -> unit
\end{lstlisting}
Note that we are able to use \lstinline[style=oxcaml]{reify} rather to
than \lstinline[style=oxcaml]{reify_global} allowing the function passed
to \lstinline[style=oxcaml]{handle_state} to be local.

This shows how to implement state using effect reflection, but the same
interface can also be implemented directly in OxCaml using mutable state
and locality:
\begin{lstlisting}[style=oxcaml]
    type handler = int ref

    let handle_state i f = f (ref i)
    let get r = !r
    let set r i = r := i
\end{lstlisting}
This implementation gives the same behaviour as the one built using
effect reflection. Crucially it is just as easy to reason about, which
in turn means it is just as easy to reason about as the state
monad. This shows that the improved reasoning ability of the state monad
over ordinary mutable state is precisely captured by the notion of
locality.

This transformation from effect reflection to local mutable state works
for any \emph{tail-resumptive}\cite{xie2020effect} effect handler.

\subsection{Example: Generators}

A more realistic use of effect reflection is a \emph{generator}
effect (Fig.~\ref{ex-generator}).

\begin{figure}
\begin{lstlisting}[style=oxcaml]
  module Gen = struct
    type 'a t =
      | Gen : int -> unit t
  end

  module Gen_eff = Reflection(Gen)
  open Term(Gen)

  type ints =
    | Finished
    | More of int * (unit -> ints)

  let handle_gen f =
    let rec loop = function
      | Return () -> Finished
      | Perform(Gen i, k) -> More(i, fun () -> loop (Gen_eff.continue_global k ()))
    in
    loop (Gen_eff.reify_global f)

  let gen h i =
    Gen_eff.perform (Gen i) h
\end{lstlisting}
\caption{Example: Generators}
\label{fig:ex-generator}
\end{figure}

This effect has a single operation \lstinline[style=oxcaml]{Gen}, along
with a handler that turns computations using
\lstinline[style=oxcaml]{Gen} into streams of integers.

Unlike the state handler, here we must use
\lstinline[style=oxcaml]{reify_global} rather than
\lstinline[style=oxcaml]{reify} because our intention is that the
resulting stream of integers be a global value that can escape the
current region. Since it closes over the continuation
\lstinline[style=oxcaml]{k}, we must use \lstinline[style=oxcaml]{reify_global}
and require that \lstinline[style=oxcaml]{f} be global:
\begin{lstlisting}[style=oxcaml]
  val handle_gen : (Handler(Gen).t @ local -> unit) -> ints
\end{lstlisting}

\subsection{Example: Asynchronous I/O}

Perhaps the most common use case for algebraic effects is for
concurrency and asynchronous I/O. Implementing asynchronous I/O from
scratch is beyond the scope of this paper, but we can build a simple
version on top of an existing monadic concurrency library
(Fig.~\ref{fig:ex-aio}). Assuming we have a monadic future type
\lstinline[style=oxcaml]{'a Deferred.t} we can write concurrent code in
direct style by using an effect with a single
\lstinline[style=oxcaml]{Await} operation that waits for a future to be
completed.

\begin{figure}
\begin{lstlisting}[style=oxcaml]
  module Await : sig
    type 'a t =
      | Await : 'a Deferred.t -> 'a t
  end

  module Await_eff = Reflection(Await)
  open Term(Await)

  let handle_await f =
    let rec loop = function
      | Return x -> Deferred.return x
      | Perform(Await d, k) ->
          Deferred.bind d (fun x -> loop (Await_eff.continue_global k x))
    in
    loop (Await_eff.reify_global f)

  let await h d =
    Await_eff.perform (Await d) h
\end{lstlisting}
\caption{Example: Asynchronous I/O}
\label{fig:ex-aio}
\end{figure}

Using this we can write direct-style code using asynchronous I/O:
\begin{lstlisting}[style=oxcaml]
    let copy_file h src dst =
      let s = await h (File.read src) in
      await h (File.write dst s)
\end{lstlisting}

\subsection{Nested effect handlers}

A key feature of algebraic effects is how easily they compose. We can
use state whilst using asynchronous I/O by simply nesting their
handlers:
\begin{lstlisting}[style=oxcaml]
    handle_await (fun ah ->
      handle_state i (fun sh ->
        List.iter (fun fl ->
          let s = await ah (File.read fl) in
          let i = int_of_string s in
          set sh (i + get sh)) files;
        get sh))
\end{lstlisting}

This relies on the fact that \lstinline[style=oxcaml]{handle_state}
uses \lstinline[style=oxcaml]{reify} rather than
\lstinline[style=oxcaml]{reify_global} and so is able to operate on a
local computation that closes over the local await handler value
\lstinline[style=oxcaml]{ah}.

If we try to do something similar using our
\lstinline[style=oxcaml]{handle_gen} handler, then we get an error:
\begin{lstlisting}[style=oxcaml]
    handle_await (fun ah ->
      handle_gen (fun gh ->
        List.iter (fun fl ->
          let s = await ah (File.read fl) in
          let i = int_of_string s in
          gen gh i) files))
    (*@ \textcolor{red}{Error: let s = await ah (File.read fl) in} @*)
                         (*@ \textcolor{red}{\textasciicircum\textasciicircum} @*)
    (*@ \textcolor{red}{ This value is local and would escape its region.} @*)
\end{lstlisting}

We could change the definition of \lstinline[style=oxcaml]{handle_gen}
to use \lstinline[style=oxcaml]{reify} instead of
\lstinline[style=oxcaml]{reify_global}, but then we would lose the property
that the generator we produce can be passed around freely. Instead we
need to build a handler that can \emph{forward} the
\lstinline[style=oxcaml]{Await} operation to different effect handlers
at different points in the program's execution.

\subsection{Effect sums and effect forwarding}

Composing handlers like \lstinline[style=oxcaml]{handle_gen}, that use
\lstinline[style=oxcaml]{reify_global} rather than
\lstinline[style=oxcaml]{reify}, is a fundamentally more involved
operation. Consider the case of generators that are able to use
asynchronous I/O. We are given a computation that performs both
\lstinline[style=oxcaml]{Await} and \lstinline[style=oxcaml]{Gen}
operations and we'd like to turn it into a stream of integers.  We would
like to be able to freely pass this stream around. In particular, we do
not wish the stream to be prevented from escaping the scope of the
current \lstinline[style=oxcaml]{Await} handler.  However forcing parts
of the stream requires performing additional
\lstinline[style=oxcaml]{Await} operations, so whenever we force part of
the stream we'll need to provide a handler for those
\lstinline[style=oxcaml]{Await} operations again.

To achieve this more general form of composition, we must use the
selection operations \lstinline[style=oxcaml]{outl} and
\lstinline[style=oxcaml]{outr}. Using these we can write our desired
handler for the \lstinline[style=oxcaml]{Gen} effect by instead writing
a handler for \lstinline[style=oxcaml]{Sum(Await)(Gen)} that forwards
all the \lstinline[style=oxcaml]{Await} operations on to another
handler.

\begin{lstlisting}[style=oxcaml]
  module Eff = Reflection(Sum(Await)(Gen))
  module Sel = Select(Await)(Gen)

  type aints =
    | Finished
    | More of int * (Handler(Await).t @ local -> aints)

  let handle_gen_in_await f ah =
    let rec loop ah = function
      | Return () -> Finished
      | Perform(Left op, k) -> loop ah (Eff.continue_global k (perform ah op))
      | Perform(Right(Gen i), k) ->
          More(i, fun ah -> loop ah (Eff.continue_global k ()))
    in
    loop ah (Eff.reify_global (fun h -> f (Sel.outl h) (Sel.outr h))
\end{lstlisting}

The need for explicit effect forwarding is a short-coming of all systems
based around effect capabilities. This includes named handlers
(\cite{xie2022first}) and lexical effect handlers
(\cite{biernacki2019binders}). Examples like this one fundamentally
require that performing an \lstinline[style=oxcaml]{Await} operation
does work proportional to the number of handlers it is forwarded
through. Effect capabilities do a constant amount of work to perform
operations and so cannot handle such examples directly. Traditional
effect handlers, on the other hand, \textbf{always} do work proportional
to the number of nested handlers, allowing them to support such examples
directly but paying a higher cost for the cases that do not need
forwarding.

\end{document}
